---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:
-   Ensure data quality
-   Identify key predictors
-   Guide model choice and feature engineering


## Setting the environment up

### Loading packages to use

```{r warning = FALSE, message = FALSE}
## Custom functions
library('project.nyc.taxi')

## To manage relative paths
library(here)

## To transform data larger than RAM
library(DBI)
library(duckdb)

## To transform data that fits in RAM
library(data.table)
library(lubridate)
library(future)
library(future.apply)

## To create plots
library(ggplot2)
library(scales)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Creating DB connections

```{r}
con <- dbConnect(duckdb(), dbdir = here("my-db.duckdb"))

dbListTables(conn = con)
```

### Sampling data to use

As we have too much data will only need 1% of the data for training and testing the model, which corresponds to 3,272,947 rows between 2022 and 2023.

```r
ValidZoneSampleQuery <- glue::glue("
CREATE TABLE ValidZoneSample AS
SELECT t1.*
FROM NycTrips t1
INNER JOIN ZoneCodesFilter t2
  ON t1.PULocationID = t2.PULocationID AND
     t1.DOLocationID = t2.DOLocationID
USING SAMPLE 1% (system, 547548);
")

dbExecute(con, ValidZoneSampleQuery)
```

### Adding the variable to predict

Once we have a sample of the data, we can add the variable to predict as it isn't part of the of original data by following the next steps:

1. Importing the data.

```{r}
PointMeanDistance <- dbGetQuery(con, "SELECT * FROM PointMeanDistance")
ValidZoneSample <- dbGetQuery(con, "SELECT * FROM ValidZoneSample")
```

2. Changing tables to become data.tables.

```{r}
setDT(ValidZoneSample)
setDT(PointMeanDistance)
```

3. Split the data by month and link with the original parquet files.

```{r}
ValidZoneSampleByMonth <-
  ValidZoneSample[
    j = request_datetime_extra := request_datetime + minutes(15)
  ][floor_date(request_datetime_extra, unit = "month") == floor_date(request_datetime, unit = "month"),
    .(data = list(.SD)), 
    keyby = c("year", "month"),
    .SDcols = !c("request_datetime_extra")
  ][, source_path := dir("raw-data/trip-data", recursive = TRUE, full.names = TRUE)[1]]

tibble::as_tibble(ValidZoneSampleByMonth)
```

3. Run the process in paralell

```{r}
data.table::setDTthreads(1)
options(future.globals.maxSize = 10 * 1e9)
plan(multicore, workers = 7)

profvis::profvis({
ValidZoneSampleByMonth[, add_take_current_trip(trip_sample = data[[1L]],
                                               point_mean_distance = PointMeanDistance,
                                               parquet_path = source_path,
                                               future.scheduling = 0.5,
                                               future.chunk.size = 100),
                       by = c("year", "month")]
})

```

```{r}
dbDisconnect(con, shutdown = TRUE)

rm(con)
```

