---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:
-   Ensure data quality
-   Identify key predictors
-   Detect multicollinearity
-   Guide model choice and feature engineering


## Setting the environment up

### Loading packages to use

```{r warning = FALSE, message = FALSE}
## Custom functions
library('project.nyc.taxi')

## To manage relative paths
library(here)

## To transform data larger than RAM
library(DBI)
library(duckdb)

## To transform data that fits in RAM
library(data.table)
library(lubridate)
library(future)
library(future.apply)

## To create plots
library(ggplot2)
library(scales)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Creating DB connections

```{r}
con <- dbConnect(duckdb(), dbdir = here("my-db.duckdb"))

dbListTables(conn = con)
```

### Sampling data to use

As we have too much data will only need 1% of the data for training and testing the model, which corresponds to 3,272,947 rows between 2022 and 2023.

```r
ValidZoneSampleQuery <- glue::glue("
CREATE TABLE ValidZoneSample AS
SELECT t1.*
FROM NycTrips t1
INNER JOIN ZoneCodesFilter t2
  ON t1.PULocationID = t2.PULocationID AND
     t1.DOLocationID = t2.DOLocationID
USING SAMPLE 1% (system, 547548);
")

dbExecute(con, ValidZoneSampleQuery)
```

### Evaluate the performance

```{r}
ValidZoneSample <- dbGetQuery(con, "SELECT * FROM ValidZoneSample ORDER BY year, month")

setDT(ValidZoneSample)

ValidZoneSample =
  ValidZoneSample[, request_datetime_extra := request_datetime + minutes(15)
  # For computational limitations we will 
  # exclude trips that need to search in different months
  ][floor_date(request_datetime_extra, unit = "month") == 
      floor_date(request_datetime, unit = "month")]

SampleMonths <- ValidZoneSample[, .N, keyby = c("year", "month")]
```

```{r}
PointMeanDistance <- dbGetQuery(con, "SELECT * FROM PointMeanDistance")

setDT(PointMeanDistance)
```

```{r}
ParquetFiles <-
  here::here("raw-data", "trip-data") |>
  list.files(recursive = TRUE,
             full.names = TRUE) |>
  paste0("'", a = _ ,"'") |> 
  paste0("read_parquet([", a = _ ,"])")
```


```r
options(future.globals.maxSize = 7 * 1024^3)
plan(multisession, workers = 4)

results <-
  lapply(1:nrow(SampleMonths), \(month_i){
    
    ValidZoneSample_i <- ValidZoneSample[SampleMonths[month_i], on = c("year", "month")]
    NycTrips_i <- dbGetQuery(con,glue("SELECT * FROM {ParquetFiles[month_i]}"))
    setDT(NycTrips_i)
    
    
    # Step 1: Calculate waiting_secs and performance_per_hour for joined data
    trip_data <-
      future_lapply(1:nrow(ValidZoneSample_i), \(n_row){
        
        time_trips <-
          NycTrips_i[ValidZoneSample_i[n_row], 
                     on = "hvfhs_license_num"
          ][request_datetime >= i.request_datetime &
              request_datetime <= request_datetime_extra]
        
        current_trip_alternatives <-
          PointMeanDistance[.(ValidZoneSample_i$`PULocationID`[n_row]),
                            on = "PULocationID",
                            nomatch = NULL
          ][time_trips,
            on = c("DOLocationID" = "PULocationID"),
          ][, waiting_secs := as.numeric(difftime(request_datetime, i.request_datetime, units = "secs"))
          ][, performance_per_hour := (driver_pay + tips) / ((trip_time + waiting_secs) / 3600)
          ][(request_datetime <= i.request_datetime + 60 & trip_miles_mean <= 1) |
              (request_datetime <= i.request_datetime + 180 & trip_miles_mean <= 3) |
              (request_datetime <= i.request_datetime + 300 & trip_miles_mean <= 5) |
              (request_datetime <= i.request_datetime + 420 & trip_miles_mean <= 7) |
              (request_datetime <= i.request_datetime + 540 & trip_miles_mean <= 9) |
              (request_datetime <= i.request_datetime + 660 & trip_miles_mean <= 11) |
              (request_datetime <= i.request_datetime + 780 & trip_miles_mean <= 13) |
              (request_datetime <= i.request_datetime + 900 & trip_miles_mean <= 15) &
              ((i.wav_match_flag == 'Y' & wav_match_flag %in% c('Y', 'N')) |
                 (i.wav_match_flag == 'N' & wav_match_flag == 'N'))]
        
        return(current_trip_alternatives)
        
      }) |>
      rbindlist()
    
    # Step 2: Aggregate data by performance metrics
    aggregated_data <- trip_data[
      , .(percentile_75_performance = quantile(performance_per_hour, 0.75, na.rm = TRUE)),
      by = "trip_id"
    ]
    
    
    # Step 3: Join aggregated data with ValidZoneSample
    final_data <- 
      ValidZoneSample_i[aggregated_data, 
                        on = "trip_id"
      ][, take_current_trip := fifelse(performance_per_hour > percentile_75_performance, 0L, 1L)]
    
    
    # Result
    return(final_data[])
    
  })|>
  rbindlist()
```

```{r}
dbDisconnect(con, shutdown = TRUE)

rm(con)
```

