---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:

-   Ensure data quality

-   Identify key predictors

-   Guide model choice and feature engineering

## Setting the environment up

### Define colors to use

```{r}
ColorHighlight <- "lightslateblue"
ColorGray <- "gray80"
```


### Loading packages to use

```{r warning = FALSE, message = FALSE}
## To manage relative paths
library(here)

## To import and export data frames as binary files
library(fst)
library(fstcore)

## To transform data that fits in RAM
library(data.table)
library(lubridate)

## To create plots
library(ggplot2)
library(scales)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Importing training sample

```{r}
TrainingSample <-
  here("output/take-trip-fst") |>
  list.files(full.names = TRUE) |>
  (\(x) data.table(full_path = x,
                   n_char = nchar(basename(x)),
                   name = basename(x)))() |>
  (\(dt) dt[order(n_char, name), full_path])() |>
  head(12L) |>
  lapply(FUN = read_fst, as.data.table = TRUE) |>
  rbindlist()

## Changing target as character
TrainingSample[, take_current_trip := fifelse(take_current_trip == 1, "Y", "N")]
```

## Explorting individual distributions

### Categorical  variables

These plots confirm that:

- Uber accounts for 72% of trips. In consequence, the results of the this project will be __more related to Uber than Lyft__.

- `dispatching_base_num`, `originating_base_num` and `access_a_ride_flag` all seem to be highly correlated with company and are __unlikely to be useful for prediction__.

- Shared rides and WAV (Wheelchair Accessible Vehicle) rides are uncommon.

- If you request a shared trip, you will only be matched with other riders in 23% of cases.

- When someone requests a WAV, they are able to find this type of vehicle most of the time.

- As most trips can be taken (76%), we might need to __over-sample the trips that shouldn't be taken__ to improve model performance for a class imbalance.

::: {.panel-tabset}

#### Company

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = .(company = fifelse(hvfhs_license_num == "HV0003", "Uber", "Lyft"))
][, prop := N / sum(N)] |>
  ggplot(aes(N, company)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips comes from Uber",
       x = "",
       y = "Company Name") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### dispatching_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("dispatching_base_num", "hvfhs_license_num")
][,`:=`(prop  = N / sum(N),
        dispatching_base_num = reorder(dispatching_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, dispatching_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### originating_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "originating_base_num")
][,`:=`(prop  = N / sum(N),
        originating_base_num = reorder(originating_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, originating_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### shared_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "shared_request_flag"
][,`:=`(prop = N / sum(N),
        shared_request_flag = reorder(shared_request_flag, N))] |>
  ggplot(aes(N, shared_request_flag)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 8000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Shared trips are weird") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### shared_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("shared_request_flag",
                      "shared_match_flag")
][,prop := N / sum(N),
  by = "shared_request_flag"] |>
  ggplot(aes(shared_match_flag,
             shared_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = percent)+
  scale_x_discrete(position = "top")+
  labs(title = "It's to match a shared trip",
       fill = "Based on request proportion") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 

```

#### access_a_ride_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "access_a_ride_flag")
][,`:=`(prop  = N / sum(N),
        access_a_ride_flag = reorder(access_a_ride_flag, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, access_a_ride_flag)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "Don't  show any new detail",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### wav_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "wav_request_flag"
][,`:=`(prop = N / sum(N),
        wav_request_flag = reorder(wav_request_flag, N))] |>
  ggplot(aes(N, wav_request_flag)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 0.01),
                x = N + c(300000, 250))) +
  scale_x_continuous(labels = comma_format(),
                     trans = "log2",
                     breaks = 2^(seq(0, 20, by = 2))) +
  labs(title = "Really few trips need WAV",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### wav_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("wav_match_flag", "wav_request_flag")
][, prop := N / sum(N),
  by = "wav_request_flag"] |>
  ggplot(aes(wav_match_flag, wav_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "The WAV request is met",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### take_current_trip

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "take_current_trip"
][,`:=`(prop = N / sum(N),
        take_current_trip = reorder(take_current_trip, N))] |>
  ggplot(aes(N, take_current_trip)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            nudge_x = 10000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most trips can be taken",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

:::


### Datetime variables

As request `request_datetime` is the only variable we will use for predicting, we need to extract from it some feature to confirm if they can be useful.

```{r}
RequestTimeDetails <- TrainingSample[
  j = .(request_datetime,
        request_date = as_date(request_datetime),
        weekday = wday(request_datetime, label = TRUE, abbr = FALSE, week_start = 1),
        month = month(request_datetime, label = TRUE, abbr = FALSE),
        hour = hour(request_datetime))
]
```

The following can be confirmed based on the results below:

1. Trip counts show variability, with most days falling between __1,000–3,000 trips__.

2. __Strong monthly trends are present__, with peaks in specific months (February, August, September, and October).

3. Bi-weekly peaks suggest __periodic patterns__, indicating the need for temporal features like week number or bi-weekly intervals to capture this behavior.

4. __Thursday__ has significantly more trips compared to other days, revealing a clear __weekly trend__.

6. Demand consistently __increases around 12 PM__ across different days, with some variation by weekday and weekend. Incorporating hour-of-day as a feature can capture __intraday patterns__.

In summary, the data indicates strong temporal patterns so extracting features like weekday, month, hour of day, and potentially even a __sin/cosine__ transformation (for capturing seasonality) would be beneficial for model performance.

::: {.panel-tabset}

#### Bi-weekly

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(RequestTimeDetails, aes(request_datetime)) +
  geom_histogram(binwidth = 3600*24*7*2)+
  scale_x_datetime(date_breaks = "2 weeks",
                   date_labels = "%m-%d")+
  scale_y_continuous(breaks = breaks_width(5000),
                     labels = comma_format(accuracy = 1))+
  labs(title = "Distribution of BI-weekly trip requets") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Trips per day

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TripsPerDay <- RequestTimeDetails[, .N, request_date]

ggplot(TripsPerDay, aes(N)) +
  geom_histogram(bins = 15) +
  scale_x_continuous(breaks = breaks_width(1000),
                     labels = \(x) comma(x/1000, accuracy = 0.1, suffix = "k")) +
    labs(title = paste("Number of trips per day in", nrow(TripsPerDay), "dates")) +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Month

```{r}
#| code-fold: true
#| code-summary: "Show the code"

RequestTimeDetails[ 
  j = .(is_top_4 = FALSE,
        n_trips = .N),
  by = "month"
][order(-n_trips)[1:4], 
  is_top_4 := TRUE] |>
  ggplot(aes(n_trips, factor(month, sort(month, decreasing = TRUE)))) +
  geom_col(aes(fill = is_top_4),
           color = "black",
           width = 0.8) +
  scale_fill_manual(values = c("TRUE" = ColorHighlight, 
                               "FALSE" = ColorGray))+
  scale_x_continuous(labels = comma_format()) +
  labs(title = "4 months have around 30k trips",
       y = "",
       x = "Number of Trips") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Day of week

```{r}
#| code-fold: true
#| code-summary: "Show the code"

RequestTimeDetails[ 
  j = .(n_trips = .N),
  keyby = .(weekday = forcats::fct_rev(weekday))
][, is_top := n_trips == max(n_trips)] |>
  ggplot(aes(n_trips, weekday)) +
  geom_col(aes(fill = is_top),
           color = "black",
           width = 0.7) +
  scale_fill_manual(values = c("TRUE" = ColorHighlight, 
                               "FALSE" = ColorGray))+
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Must of the samples take place on Thursday",
       y = "",
       x = "Number of Trips") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Weekday by Hour

```{r}
#| code-fold: true
#| code-summary: "Show the code"

HourWeekdayPlot <-
  RequestTimeDetails[
    j = .(n_trips = .N),
    by = c("hour", "weekday")
  ][, `:=`(trip_cat = fifelse(n_trips %in% sort(n_trips, TRUE)[1:3],
                              "extreme_value",
                              "normal"),
           prop_trips = n_trips / sum(n_trips)),
    by = "weekday"] |>
  ggplot(aes(hour, 
             prop_trips, 
             colour = trip_cat)) +
  geom_point()+
  scale_color_manual(values = c("extreme_value" = ColorHighlight,
                                "normal" = ColorGray)) +
  scale_x_continuous(breaks = breaks_width(6)) +
  scale_y_continuous(labels = percent_format())+
  facet_wrap(vars(weekday)) +
  expand_limits(y = 0, x = c(0,24)) +
  labs(title = "Higher demand hours start at 12 pm",
       y = "Proportion of trips per weekday",
       x = "Hour") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))

plotly::ggplotly(HourWeekdayPlot)
```

:::

### Geospatial data


```{r}
ColsToImport <- c("LocationID", "Borough", "service_zone", "lat", "long")

ZoneCodesRefClean <-
  here("output/cache-data/ZoneCodesRefClean.fst") |>
  fst::read_fst(as.data.table = TRUE,
                columns = ColsToImport)

StartZoneCode <- copy(ZoneCodesRefClean)
EndZoneCode <- copy(ZoneCodesRefClean)

setnames(StartZoneCode, c("PULocationID", paste0("PU_", ColsToImport[-1L])))
setnames(EndZoneCode, c("DOLocationID", paste0("DO_", ColsToImport[-1L])))

GeoSpatialDetails <- TrainingSample[
  .SDcols = c("PULocationID", "DOLocationID", "trip_miles"),
  j = EndZoneCode[.SD, on = .NATURAL]
][
  j = StartZoneCode[.SD, on = .NATURAL]
]
```


```{r}
library(sf)

GeoSpatialDetails[, .(start_n = .N), 
                  .(lat = PU_lat,
                    long = PU_long)
][GeoSpatialDetails[, .(end_n = .N), 
                    .(lat = DO_lat,
                      long = DO_long)],
  on = c("lat", "long")
][, ratio := fcoalesce(start_n, 0L) / fcoalesce(end_n, 0L)][]
```




**1. Spatial Distribution of Pickups and Dropoffs**

   *   **What:** Visualize the locations of pickups (PULocationID) and dropoffs (DOLocationID) on a map. You can plot points, heatmaps, or even create kernel density estimates.
   *   **Why:**
        *   **Identify High-Demand Areas:** See which areas are most frequently used for pickups and dropoffs. This helps identify areas with high taxi demand and allows for better regionalized feature engineering.
        *   **Understand Trip Patterns:** Identify common origin-destination patterns. Are there clusters or corridors of activity?
        *   **Spot Anomalies:** Detect unusual pickup or dropoff locations that might indicate data errors or outlier trips.
   *   **How:**
        *   Use a mapping library like `ggplot2` or `leaflet` in R.
        *   Consider using `sf` objects for spatial data handling in R.
        *   Use heatmaps to show density of trips by location.

**2. Geographic Range and Coverage**

    *   **What:** Explore the geographical boundaries of the available data. Do the pickup and dropoff locations cover all of Manhattan, Brooklyn, and Queens as intended?
    *   **Why:**
        *   **Verify Scope:** Ensure that the data fits within your project scope (i.e. the three boroughs).
        *   **Edge Effects:** Understand if there are any trips occurring at the periphery of the area you intend to analyse.
        *   **Data Gaps:** Detect any spatial gaps or areas with no or very little data.
   *   **How:**
        *   Find the minimum and maximum `lat` and `long` to know how spread out the trips are.
        *   Plot the boundary using a convex hull of the points on the map.
        *   Calculate the density of trips using a kernel density function and visualize the area that contains most of the trips.

**3. Relationship between Location and Other Variables**

    *   **What:** Explore the relationship of location (lat, long) with key variables from your original dataset, like fares, trip time, driver pay, or tips.
    *   **Why:**
        *   **Detect Patterns:** See if there are regions where trips tend to be more profitable or have longer duration.
        *   **Inform Feature Engineering:** This can give hints as to what kind of geospatial features might be valuable to engineer in the following steps.
        *   **Prioritize Spatial Areas:** It might be the case that only specific areas are associated to longer trips with higher revenue.
   *   **How:**
        *   Group trips by spatial zones (e.g. using a grid or the `LocationID` values you have).
        *   Calculate the average of numerical features like `trip_miles`, `trip_time`, `driver_pay` or `tips` for each spatial group.
        *   Visualize the averages as a choropleth map using different ranges of the values.

**4. Spatial Outliers**

    *   **What:** Explore locations that are far away from most of the other trips and identify what they are.
    *   **Why:**
         *  **Detect Errors:** They might be due to data entry errors.
         *  **Highlight Unusual Trip Patterns:** If not errors, they could indicate special trips of interest.
   *   **How:**
        *   Using a box-plot or other visualization techniques, detect outliers based on their latitude or longitude, or based on the density of trips at that area.
        *   Manually check the map to see where those trips are taken.

**5. Initial Explorations with Zone Information**

    *   **What:** Map and analyze the zones you are working with using the `ZoneCodesRef` table you created. Check if the zones make sense for the kind of data you have and the area they represent.
    *   **Why:**
        *   **Understand spatial groupings:** This initial understanding will inform the kind of features you will engineer in the following step.
        *   **Spot inconsistencies:** Determine if any of the geographic locations are misaligned on your map.
   *   **How:**
        *  Use the `plot_map` function you already have defined.
        * Check that all locations make sense for the dataset.

**6. Exploring Distances:**

    *   **What:** Plot the distribution of distances between pick-up and drop-off points to see what is the pattern in the dataset.
    *  **Why:**
        *   **Define Thresholds**: It will allow to define threshold to search for data in the `add_take_current_trip` functions that is used later.
        *  **Understand Distribution:** Understand what type of spatial ranges is the most common in the dataset to inform the feature engineering.
    * **How:**
        *  Using `geosphere` package in R, calculate the distance between pick-up and drop-off points.
        * Plot the distribution using a histogram.

**How to present this information:**

   *   **Maps:** Use maps as much as possible. Spatial analysis is fundamentally visual.
   *   **Summary Statistics:** Calculate and display means, medians, standard deviations, etc. for different zones.
   *   **Histograms/Boxplots:** Show the distribution of trips and fares across spatial contexts.
   *  **Scatterplots**: See if there are relationship between locations and other variables of interest.

**In Summary, in the Data Understanding phase:**

*   **Don't engineer, analyze:** Your goal is to *analyze* the spatial data you have, not to start *creating* features yet.
*   **Visualize to understand:** Use maps to see spatial patterns in pickups and dropoffs and explore geographic outliers.
*   **Explore and Connect:** Look for correlations between location and existing variables, identifying key insights that will inform the next steps.

This should give you a solid understanding of the spatial aspect of your data, which will guide you when it comes to engineering your spatial features.



### Numerical variables

### Correlations between predictors and the target variable
