---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:

-   Ensure data quality

-   Identify key predictors

-   Guide model choice and feature engineering

## Setting the environment up

### Loading packages to use

```{r warning = FALSE, message = FALSE}
## To manage relative paths
library(here)

## To import and export data frames as binary files
library(fst)

## To transform data that fits in RAM
library(data.table)
library(lubridate)

## To create plots
library(ggplot2)
library(scales)
library(DataExplorer)
library(GGally)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Importing training sample

```{r}
TrainingSample <-
  here("output/take-trip-fst") |>
  list.files(full.names = TRUE) |>
  (\(x) data.table(full_path = x,
                   n_char = nchar(basename(x)),
                   name = basename(x)))() |>
  (\(dt) dt[order(n_char, name), full_path])() |>
  head(12L) |>
  lapply(FUN = fst::read_fst, as.data.table = TRUE) |>
  rbindlist()
```

### Formatting for EDA

https://boxuancui.github.io/DataExplorer/articles/dataexplorer-intro.html#distributions

https://ggobi.github.io/ggally/articles/ggpairs.html

```{r}
TrainingSample[, take_current_trip := fifelse(take_current_trip == 1, "Y", "N")]
```


## Explorting individual distributions

### Categorical  variables

In the next plots we can confirm that:

- __Uber__ makes 72% of the trips.
- __dispatching_base_num__, __originating_base_num__ and __access_a_ride_flag__ are highly correlated with company, in consequence we cannot use it for predicting.
- Only 1% of trips are shared and 6% needs WAV.
- __wav_match_flag__ and __wav_request_flag__ are highly correlated.

::: {.panel-tabset}

#### Company

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = .(company = fifelse(hvfhs_license_num == "HV0003", "Uber", "Lyft"))
][, prop := N / sum(N)] |>
  ggplot(aes(N, company)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips comes from Uber",
       x = "",
       y = "Company Name") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### dispatching_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("dispatching_base_num", "hvfhs_license_num")
][,`:=`(prop  = N / sum(N),
        dispatching_base_num = reorder(dispatching_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, dispatching_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### originating_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "originating_base_num")
][,`:=`(prop  = N / sum(N),
        originating_base_num = reorder(originating_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, originating_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### shared_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "shared_request_flag"
][,`:=`(prop = N / sum(N),
        shared_request_flag = reorder(shared_request_flag, N))] |>
  ggplot(aes(N, shared_request_flag)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 8000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Shared trips are weird") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### shared_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("shared_request_flag",
                      "shared_match_flag")
][,prop := N / sum(N),
  by = "shared_request_flag"] |>
  ggplot(aes(shared_match_flag,
             shared_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = percent)+
  scale_x_discrete(position = "top")+
  labs(title = "It's to match a shared trip",
       fill = "Based on request proportion") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
                                                                
```

#### access_a_ride_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "access_a_ride_flag")
][,`:=`(prop  = N / sum(N),
        access_a_ride_flag = reorder(access_a_ride_flag, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, access_a_ride_flag)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "Don't  show any new detail",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### wav_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "wav_request_flag"
][,`:=`(prop = N / sum(N),
        wav_request_flag = reorder(wav_request_flag, N))] |>
ggplot(aes(N, wav_request_flag)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 0.01),
                x = N + c(300000, 250))) +
  scale_x_continuous(labels = comma_format(),
                     trans = "log2",
                     breaks = 2^(seq(0, 20, by = 2))) +
  labs(title = "Really few trips need WAV",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### wav_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("wav_match_flag", "wav_request_flag")
][, prop := N / sum(N),
  by = "wav_request_flag"] |>
  ggplot(aes(wav_match_flag, wav_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "Don't  show any new detail",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

:::
