---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:

-   Ensure data quality

-   Identify key predictors

-   Guide model choice and feature engineering

## Setting the environment up

### Define colors to use

```{r}
ColorHighlight <- "lightslateblue"
ColorGray <- "gray80"
```


### Loading packages to use

```{r warning = FALSE, message = FALSE}
## To manage relative paths
library(here)

## To import and export data frames as binary files
library(fst)
library(fstcore)

## To transform data that fits in RAM
library(data.table)
library(lubridate)

## To create plots
library(ggplot2)
library(scales)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Importing training sample

```{r}
TrainingSample <-
  here("output/take-trip-fst") |>
  list.files(full.names = TRUE) |>
  (\(x) data.table(full_path = x,
                   n_char = nchar(basename(x)),
                   name = basename(x)))() |>
  (\(dt) dt[order(n_char, name), full_path])() |>
  head(12L) |>
  lapply(FUN = read_fst, as.data.table = TRUE) |>
  rbindlist()

## Changing target as character
TrainingSample[, take_current_trip := fifelse(take_current_trip == 1, "Y", "N")]
```

## Explorting individual distributions

### Categorical  variables

These plots confirm that:

- Uber accounts for 72% of trips. In consequence, the results of the this project will be __more related to Uber than Lyft__.

- `dispatching_base_num`, `originating_base_num` and `access_a_ride_flag` all seem to be highly correlated with company and are __unlikely to be useful for prediction__.

- Shared rides and WAV (Wheelchair Accessible Vehicle) rides are uncommon.

- If you request a shared trip, you will only be matched with other riders in 23% of cases.

- When someone requests a WAV, they are able to find this type of vehicle most of the time.

- As most trips can be taken (76%), we might need to __over-sample the trips that shouldn't be taken__ to improve model performance for a class imbalance.

::: {.panel-tabset}

#### Company

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = .(company = fifelse(hvfhs_license_num == "HV0003", "Uber", "Lyft"))
][, prop := N / sum(N)] |>
  ggplot(aes(N, company)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips comes from Uber",
       x = "",
       y = "Company Name") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### dispatching_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("dispatching_base_num", "hvfhs_license_num")
][,`:=`(prop  = N / sum(N),
        dispatching_base_num = reorder(dispatching_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, dispatching_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### originating_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "originating_base_num")
][,`:=`(prop  = N / sum(N),
        originating_base_num = reorder(originating_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, originating_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### shared_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "shared_request_flag"
][,`:=`(prop = N / sum(N),
        shared_request_flag = reorder(shared_request_flag, N))] |>
  ggplot(aes(N, shared_request_flag)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 8000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Shared trips are weird") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### shared_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("shared_request_flag",
                      "shared_match_flag")
][,prop := N / sum(N),
  by = "shared_request_flag"] |>
  ggplot(aes(shared_match_flag,
             shared_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = percent)+
  scale_x_discrete(position = "top")+
  labs(title = "It's to match a shared trip",
       fill = "Based on request proportion") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 

```

#### access_a_ride_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "access_a_ride_flag")
][,`:=`(prop  = N / sum(N),
        access_a_ride_flag = reorder(access_a_ride_flag, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, access_a_ride_flag)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "Don't  show any new detail",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### wav_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "wav_request_flag"
][,`:=`(prop = N / sum(N),
        wav_request_flag = reorder(wav_request_flag, N))] |>
  ggplot(aes(N, wav_request_flag)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 0.01),
                x = N + c(300000, 250))) +
  scale_x_continuous(labels = comma_format(),
                     trans = "log2",
                     breaks = 2^(seq(0, 20, by = 2))) +
  labs(title = "Really few trips need WAV",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### wav_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("wav_match_flag", "wav_request_flag")
][, prop := N / sum(N),
  by = "wav_request_flag"] |>
  ggplot(aes(wav_match_flag, wav_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = ColorGray, 
                      high = ColorHighlight,
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "The WAV request is met",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### take_current_trip

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "take_current_trip"
][,`:=`(prop = N / sum(N),
        take_current_trip = reorder(take_current_trip, N))] |>
  ggplot(aes(N, take_current_trip)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            nudge_x = 10000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most trips can be taken",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

:::


### Datetime variables

As request `request_datetime` is the only variable we will use for predicting, we need to extract from it some feature to confirm if they can be useful.

```{r}
RequestTimeDetails <- TrainingSample[
  j = .(request_datetime,
        request_date = as_date(request_datetime),
        weekday = wday(request_datetime, label = TRUE, abbr = FALSE, week_start = 1),
        month = month(request_datetime, label = TRUE, abbr = FALSE),
        hour = hour(request_datetime))
]
```

The following can be confirmed based on the results below:

1. Trip counts show variability, with most days falling between __1,000–3,000 trips__.

2. __Strong monthly trends are present__, with peaks in specific months (February, August, September, and October).

3. Bi-weekly peaks suggest __periodic patterns__, indicating the need for temporal features like week number or bi-weekly intervals to capture this behavior.

4. __Thursday__ has significantly more trips compared to other days, revealing a clear __weekly trend__.

6. Demand consistently __increases around 12 PM__ across different days, with some variation by weekday and weekend. Incorporating hour-of-day as a feature can capture __intraday patterns__.

In summary, the data indicates strong temporal patterns so extracting features like weekday, month, hour of day, and potentially even a __sin/cosine__ transformation (for capturing seasonality) would be beneficial for model performance.

::: {.panel-tabset}

#### Bi-weekly

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(RequestTimeDetails, aes(request_datetime)) +
  geom_histogram(binwidth = 3600*24*7*2)+
  scale_x_datetime(date_breaks = "2 weeks",
                   date_labels = "%m-%d")+
  scale_y_continuous(breaks = breaks_width(5000),
                     labels = comma_format(accuracy = 1))+
  labs(title = "Distribution of BI-weekly trip requets") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Trips per day

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TripsPerDay <- RequestTimeDetails[, .N, request_date]

ggplot(TripsPerDay, aes(N)) +
  geom_histogram(bins = 15) +
  scale_x_continuous(breaks = breaks_width(1000),
                     labels = \(x) comma(x/1000, accuracy = 0.1, suffix = "k")) +
    labs(title = paste("Number of trips per day in", nrow(TripsPerDay), "dates")) +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Month

```{r}
#| code-fold: true
#| code-summary: "Show the code"

RequestTimeDetails[ 
  j = .(is_top_4 = FALSE,
        n_trips = .N),
  by = "month"
][order(-n_trips)[1:4], 
  is_top_4 := TRUE] |>
  ggplot(aes(n_trips, factor(month, sort(month, decreasing = TRUE)))) +
  geom_col(aes(fill = is_top_4),
           color = "black",
           width = 0.8) +
  scale_fill_manual(values = c("TRUE" = ColorHighlight, 
                               "FALSE" = ColorGray))+
  scale_x_continuous(labels = comma_format()) +
  labs(title = "4 months have around 30k trips",
       y = "",
       x = "Number of Trips") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Day of week

```{r}
#| code-fold: true
#| code-summary: "Show the code"

RequestTimeDetails[ 
  j = .(n_trips = .N),
  keyby = .(weekday = forcats::fct_rev(weekday))
][, is_top := n_trips == max(n_trips)] |>
  ggplot(aes(n_trips, weekday)) +
  geom_col(aes(fill = is_top),
           color = "black",
           width = 0.7) +
  scale_fill_manual(values = c("TRUE" = ColorHighlight, 
                               "FALSE" = ColorGray))+
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Must of the samples take place on Thursday",
       y = "",
       x = "Number of Trips") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Weekday by Hour

```{r}
#| code-fold: true
#| code-summary: "Show the code"

HourWeekdayPlot <-
  RequestTimeDetails[
    j = .(n_trips = .N),
    by = c("hour", "weekday")
  ][, `:=`(trip_cat = fifelse(n_trips %in% sort(n_trips, TRUE)[1:3],
                              "extreme_value",
                              "normal"),
           prop_trips = n_trips / sum(n_trips)),
    by = "weekday"] |>
  ggplot(aes(hour, 
             prop_trips, 
             colour = trip_cat)) +
  geom_point()+
  scale_color_manual(values = c("extreme_value" = ColorHighlight,
                                "normal" = ColorGray)) +
  scale_x_continuous(breaks = breaks_width(6)) +
  scale_y_continuous(labels = percent_format())+
  facet_wrap(vars(weekday)) +
  expand_limits(y = 0, x = c(0,24)) +
  labs(title = "Higher demand hours start at 12 pm",
       y = "Proportion of trips per weekday",
       x = "Hour") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))

plotly::ggplotly(HourWeekdayPlot)
```

:::

### Geospatial data


```{r}
ZoneCodesRef <-
  fread(here("raw-data/taxi_zone_lookup.csv"),
        select = c("LocationID" = "integer",
                   "Borough" = "character",
                   "service_zone" = "character"))

TrainingSamplePickUp <-
  ZoneCodesRef[TrainingSample[, c("PULocationID",
                                  "take_current_trip")],
               on = c("LocationID" = "PULocationID")]

TrainingSampleDropOff <-
  ZoneCodesRef[TrainingSample[, c("DOLocationID",
                                  "take_current_trip")],
               on = c("LocationID" = "DOLocationID")]
```


1. Good trips take place in points with high demand.

::: {.panel-tabset}

#### Pickup Borough 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSamplePickUp[, .N,
                     by = "Borough"
][, `:=`(prop = N / sum(N),
         Borough = reorder(Borough, N))] |>
  ggplot(aes(N, Borough)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.7) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips start from Manhattan",
       x = "",
       y = "Borough") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Dropoff Borough 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSampleDropOff[, .N,
                      by = "Borough"
][, `:=`(prop = N / sum(N),
         Borough = reorder(Borough, N))] |>
  ggplot(aes(N, Borough)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.7) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips end in Manhattan",
       x = "",
       y = "Borough") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Pickup Service Zone 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSamplePickUp[, .N,
                      by = "service_zone"
][, `:=`(prop = N / sum(N),
         service_zone = reorder(service_zone, N))] |>
  ggplot(aes(N, service_zone)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.7) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips start in Boro Zone",
       x = "",
       y = "Service Zone") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Dropoff Service Zone 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSampleDropOff[, .N,
                      by = "service_zone"
][, `:=`(prop = N / sum(N),
         service_zone = reorder(service_zone, N))] |>
  ggplot(aes(N, service_zone)) +
  geom_col(fill = ColorGray,
           color = "black",
           width = 0.7) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips end in Boro Zone",
       x = "",
       y = "Service Zone") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Pickups Distribution

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSamplePickUp[, .(n_trips = .N),
                     by = "LocationID"] |>
  ggplot(aes(n_trips)) +
  geom_histogram(position = position_identity(),
                 binwidth = 200)+
  scale_y_continuous(breaks = breaks_width(5))+
  scale_x_continuous(breaks = breaks_width(500),
                     labels = comma_format())+
  labs(title = "Distribution Pickups Number of Trips",
       x = "Number of Trips",
       y = "Number of Locations",
       fill = "Take Current Trip") +
  theme_minimal()+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Dropoff Distribution

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSampleDropOff[, .(n_trips = .N),
                     by = "LocationID"] |>
  ggplot(aes(n_trips)) +
  geom_histogram(position = position_identity(),
                 binwidth = 200)+
  scale_y_continuous(breaks = breaks_width(5))+
  scale_x_continuous(breaks = breaks_width(500),
                     labels = comma_format())+
  labs(title = "Distribution Dropoffs Number of Trips",
       x = "Number of Trips",
       y = "Number of Locations",
       fill = "Take Current Trip") +
  theme_minimal()+
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

:::


### Numerical variables

### Correlations between predictors and the target variable
