---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:

-   Ensure data quality

-   Identify key predictors

-   Guide model choice and feature engineering

## Setting the environment up

### Loading packages to use

```{r warning = FALSE, message = FALSE}
## To manage relative paths
library(here)

## To import and export data frames as binary files
library(fst)
library(fstcore)

## To transform data that fits in RAM
library(data.table)
library(lubridate)
library(timeDate)
library(recipes)

## To create plots
library(ggplot2)
library(scales)
library(plotly)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Importing training sample

```{r}
TrainingSample <-
  here("output/take-trip-fst") |>
  list.files(full.names = TRUE) |>
  (\(x) data.table(full_path = x,
                   n_char = nchar(basename(x)),
                   name = basename(x)))() |>
  (\(dt) dt[order(n_char, name), full_path])() |>
  head(12L) |>
  lapply(FUN = read_fst, as.data.table = TRUE) |>
  rbindlist()

## Changing target as character
TrainingSample[, take_current_trip := fifelse(take_current_trip == 1, "Y", "N")]
```

## Explorting individual distributions

### Categorical  variables

In these plots we can confirm that:

- __Uber__ accounts for 72% of all trips.

- __dispatching_base_num__, __originating_base_num__ and __access_a_ride_flag__ are highly correlated with company and don't seem to be useful for prediction.

- Shared rides and WAV (Wheelchair Accessible Vehicle) rides are uncommon.

- If you request a shared trip, you will only be matched with other riders in 23% of cases.

- When someone requests a WAV, they are able to find this type of vehicle most of the time.

::: {.panel-tabset}

#### Company

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = .(company = fifelse(hvfhs_license_num == "HV0003", "Uber", "Lyft"))
][, prop := N / sum(N)] |>
  ggplot(aes(N, company)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 6500) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Most of the trips comes from Uber",
       x = "",
       y = "Company Name") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### dispatching_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("dispatching_base_num", "hvfhs_license_num")
][,`:=`(prop  = N / sum(N),
        dispatching_base_num = reorder(dispatching_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, dispatching_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### originating_base_num

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "originating_base_num")
][,`:=`(prop  = N / sum(N),
        originating_base_num = reorder(originating_base_num, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, originating_base_num)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "We only have 1 dispaching base per Company",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### shared_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "shared_request_flag"
][,`:=`(prop = N / sum(N),
        shared_request_flag = reorder(shared_request_flag, N))] |>
  ggplot(aes(N, shared_request_flag)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop)),
            nudge_x = 8000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Shared trips are weird") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### shared_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("shared_request_flag",
                      "shared_match_flag")
][,prop := N / sum(N),
  by = "shared_request_flag"] |>
  ggplot(aes(shared_match_flag,
             shared_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = percent)+
  scale_x_discrete(position = "top")+
  labs(title = "It's to match a shared trip",
       fill = "Based on request proportion") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 

```

#### access_a_ride_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("hvfhs_license_num",
                      "access_a_ride_flag")
][,`:=`(prop  = N / sum(N),
        access_a_ride_flag = reorder(access_a_ride_flag, N),
        hvfhs_license_num = reorder(hvfhs_license_num, N))] |>
  ggplot(aes(hvfhs_license_num, access_a_ride_flag)) +
  geom_tile(aes(fill = N))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "Don't  show any new detail",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### wav_request_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "wav_request_flag"
][,`:=`(prop = N / sum(N),
        wav_request_flag = reorder(wav_request_flag, N))] |>
  ggplot(aes(N, wav_request_flag)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 0.01),
                x = N + c(300000, 250))) +
  scale_x_continuous(labels = comma_format(),
                     trans = "log2",
                     breaks = 2^(seq(0, 20, by = 2))) +
  labs(title = "Really few trips need WAV",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### wav_match_flag

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = c("wav_match_flag", "wav_request_flag")
][, prop := N / sum(N),
  by = "wav_request_flag"] |>
  ggplot(aes(wav_match_flag, wav_request_flag)) +
  geom_tile(aes(fill = prop))+
  geom_text(aes(label = percent(prop, accuracy = 1))) +
  scale_fill_gradient(low = "grey80", 
                      high = "red",
                      labels = comma)+
  scale_x_discrete(position = "top")+
  labs(title = "The WAV request is met",
       fill = "Number of trips") +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15)) 
```

#### take_current_trip

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TrainingSample[, .N,
               by = "take_current_trip"
][,`:=`(prop = N / sum(N),
        take_current_trip = reorder(take_current_trip, N))] |>
  ggplot(aes(N, take_current_trip)) +
  geom_col(fill = "gray80",
           color = "black",
           width = 0.5) +
  geom_text(aes(label = percent(prop, accuracy = 1)),
            nudge_x = 10000) +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Must of the trips can be taken",
       x = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

:::


### Datetime variables

As request `request_datetime` is the only variable we will use for predicting, we need to extract from it some feature to cofirm if they can be useful.

```{r}
RequestTimeDetails <- TrainingSample[
  j = .(request_datetime,
        request_date = as_date(request_datetime),
        is_holiday = as_date(request_datetime) %in% as_date(holidayNYSE(2022)),
        is_weekend = isWeekend(request_datetime),
        weekday = wday(request_datetime, label = TRUE, abbr = FALSE),
        month = month(request_datetime, label = TRUE, abbr = FALSE),
        hour = hour(request_datetime))
]
```

Based on the results that we can see below, we can confirm that:

1. We are missing date for first week of July and second week of December.
2. We have dates with an unusual number of trips ending January. Taking a deeper look we can confirm that most sampled date were Jan 27, Jan 31 and Feb 05.
3. The months with higher number or trips are February, August, September and October.


::: {.panel-tabset}

#### Bi-weekly

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(RequestTimeDetails, aes(request_datetime)) +
  geom_histogram(binwidth = 3600*24*7*2)+
  scale_x_datetime(date_breaks = "2 weeks",
                   date_labels = "%m-%d")+
  scale_y_continuous(breaks = breaks_width(5000),
                     labels = comma_format(accuracy = 1))+
  labs(title = "Distribution of BI-weekly trip requets") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Trips per day

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TripsPerDay <- RequestTimeDetails[, .N, request_date]

ggplot(TripsPerDay, aes(N)) +
  geom_histogram(bins = 15) +
  scale_x_continuous(breaks = breaks_width(1000),
                     labels = \(x) comma(x/1000, accuracy = 0.1, suffix = "k")) +
    labs(title = paste("Number of trips per day in", nrow(TripsPerDay), "dates")) +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Busiest Travel Dates

```{r}
#| code-fold: true
#| code-summary: "Show the code"

TripsPerDay[order(-N)[1:10],
            .(request_date = reorder(format(request_date, "%a, %b %d"),
                                     -as.numeric(request_date)),
              N)
  ][, is_top_3 := c(rep(TRUE, 3), rep(FALSE, 7))] |>
  ggplot(aes(N, request_date, fill = is_top_3)) +
  geom_col(color = "black",
           width = 0.7) +
  scale_x_continuous(breaks = breaks_width(1000),
                     labels = comma_format()) +
  scale_fill_manual(values = c("TRUE" = "skyblue", "FALSE" = "gray80"))+
  labs(title = "Busiest Travel Dates: A Closer Look",
       x = "",
       y = "Request Date") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```

#### Month

```{r}
#| code-fold: true
#| code-summary: "Show the code"

RequestTimeDetails[ 
  j = .(is_top_4 = FALSE,
        n_trips = .N),
  by = "month"
][order(-n_trips)[1:4], 
  is_top_4 := TRUE] |>
  ggplot(aes(n_trips, factor(month, sort(month, decreasing = TRUE)))) +
  geom_col(aes(fill = is_top_4),
           color = "black",
           width = 0.8) +
  scale_fill_manual(values = c("TRUE" = "skyblue", "FALSE" = "gray80"))+
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Feb, Aug and Sept have more samples",
       y = "",
       x = "Number of Trips") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Hour

```{r}
#| code-fold: true
#| code-summary: "Show the code"

NTripsByHour <- RequestTimeDetails[
  j = .(n_trips = .N,
        is_top_hour = FALSE),
  by = hour
]

NTripsByHour[order(-n_trips)[1:2], is_top_hour := TRUE]

TopHours <- NTripsByHour[is_top_hour == TRUE, hour] |> sort()

ggplot(NTripsByHour,
       aes(n_trips, 
           factor(hour, sort(hour, decreasing = TRUE)),
           fill = is_top_hour)) +
  geom_col(color = "black",
           linewidth = 0.3) +
  scale_fill_manual(values = c("FALSE"= "grey80", "TRUE" = "skyblue")) +
  labs(title = paste0("Hightest demand at ", TopHours[1]," and ",TopHours[2]," hours"),
       y = "",
       x = "Number of Trips") +
  theme_minimal()+
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 15))
```


#### Holy days

```{r}
RequestTimeDetails[, .(avg_n_trips = .N / uniqueN(request_date)),
                   by = "is_holiday"]
```

#### Day of week

```{r}
RequestTimeDetails[, .(avg_n_trips = .N / uniqueN(request_date)),
                   by = c("is_weekend","weekday")]
```

:::


### Numeric variables

```{r}
TrainingSample[, sapply(.SD, \(x) paste0(class(x), collapse = "_") ), .SDcols = !is.character ]
```

