---
title: "Data Understanding"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After completing the [business understanding](https://angelfelizr.github.io/NycTaxi/phases/02-business-understanding.html) phase we are ready to perform the **data understanding** phase by performing an EDA with the following steps:

1.  Exploring the individual distribution of variables.
2.  Exploring correlations between predictors and target variable.
3.  Exploring correlations between predictors.

As this will help to:

-   Ensure data quality

-   Identify key predictors

-   Guide model choice and feature engineering

## Setting the environment up

### Loading packages to use

```{r warning = FALSE, message = FALSE}
## To manage relative paths
library(here)

## To import and export data frames as binary files
library(fst)

## To transform data that fits in RAM
library(data.table)
library(lubridate)

## To create plots
library(ggplot2)
library(ggiraph)

## Defining the print params to use in the report
options(datatable.print.nrows = 15, digits = 4)
```

### Importing training sample

```{r}
TrainingSample <-
  here("output/take-trip-fst") |>
  list.files(full.names = TRUE) |>
  (\(x) data.table(full_path = x,
                   n_char = nchar(basename(x)),
                   name = basename(x)))() |>
  (\(dt) dt[order(n_char, name), full_path])() |>
  head(12L) |>
  lapply(FUN = fst::read_fst, as.data.table = TRUE) |>
  rbindlist()
```


