---
title: "Expanding Geospatial Information"
editor_options: 
  chunk_output_type: console
---

## Problem description

Our target variable varies across zones. Merely providing an ID offers limited context for the models to differentiate between zones. Thus, we enrich the data with geospatial features derived from OpenStreetMap (OSM) to provide deeper insights into each zone.

## Strategy

1. **Define Spatial Extent**
   - Retrieve borough boundaries for Queens, Brooklyn, and Manhattan using `osmdata::getbb()` with polygon output.

2. **Extract OSM Features by Geometry Type**
   - **Points (e.g., amenity, shop, office, tourism, public_transport, aeroway, emergency, sport):**
     - Count occurrences within each zone.
   - **Lines (e.g., highway):**
     - Compute total road length per zone.
   - **Polygons (e.g., landuse, building):**
     - Calculate area using `st_area(geometry)`.
   - Use `add_osm_feature` for batch extraction to improve efficiency.

3. **Spatial Join with Taxi Zones**
   - Import the Taxi Zone Shapefile.
   - Join the OSM features to each zone based on spatial overlap.

4. **Feature Engineering**
   - Normalize raw counts/measurements by zone area to create density metrics (e.g., `amenity_density = n_amenities / zone_area_km2`).
   - Derive additional metrics:
     - Compute distances from zone centroids to key landmarks (e.g., airports, subway stations).
     - Determine the dominant landuse within each zone.

5. **Exploratory Data Analysis and Data Quality Checks**
   - Evaluate distributions and correlations of the new OSM features.
   - Identify and address missing values and outliers.
   - Confirm relationships between engineered features and the target variable.

6. **Modeling Preparation**
   - Merge the enriched OSM features with the training and testing datasets.


```{r}
ZonesShapes <-
  read_sf(here("raw-data/taxi_zones/taxi_zones.shp")) |>
  subset(borough %chin% c("Queens", "Brooklyn", "Manhattan"))

plot(ZonesShapes["borough"])
```

## Importing libraries

```{r}
library(here)
library(fst)
library(data.table)
library(sf)
library(tmap)
library(osmdata)

tmap_mode("view")

BoroughColors <- c(
  'Manhattan' = '#e41a1c',
  'Queens' = '#377eb8',
  'Brooklyn'= '#4daf4a'
)
```



```{r}
# Compute the overall bounding box
bb_box <-
  lapply(X = c("Manhattan, NY", "Brooklyn, NY", "Queens, NY"),
         FUN = osmdata::getbb) |>
  do.call(what = "cbind") |>
  (\(x) cbind(min = apply(x, 1, min),
              max = apply(x, 1, max)))()

project.nyc.taxi::plot_box(bb_box)
```

### Getting data


**How to Use the Features:**

-   **Radius-Based Aggregation:** Most OSM features will be aggregated within a radius around each pick-up and drop-off location. Experiment with different radius sizes (e.g., 0.5 km, 1 km, 2 km) to see which scales capture the most relevant information.
-   **Feature Engineering:** Combine different OSM features. For instance, you might calculate a ratio of restaurants to total commercial buildings.
-   **Join:** Use the `PULocationID` and `DOLocationID` to associate each set of OSM features to each trip from your dataset.

**Why These Features Will Help Your Model:**

-   **Contextual Awareness:** They provide more context about why certain locations are busy or have higher earnings per hour.
-   **Improved Accuracy:** By capturing the built environment and activity around pick-up and drop-off locations, your model will have more fine-grained and predictive features.
-   **Ability to Generalize:** OSM data is widely available and might help your model generalize better to different areas or future data.


```{r}
keep_valid_osm <- function(osm_list, key) {
  
  shape_list =
    c("osm_points",
      "osm_lines",
      "osm_polygons",
      "osm_multilines",
      "osm_multipolygons")
  
  for(shape_i in shape_list){
    
    if(is.null(osm_list[[shape_i]])){
      next
    }
    
    # Removing cases without key value
    osm_list[[shape_i]] =
      subset(osm_list[[shape_i]],
             !is.na(get(key, inherits = FALSE)),
             c("name", key))
    
    # Fixing not valied geometries
    not_valid_cases = !sf::st_is_valid(osm_list[[shape_i]])
    if(sum(not_valid_cases) > 0) {
      osm_list[[shape_i]] = sf::st_make_valid(osm_list[[shape_i]])
    }
    
  }
  
  return(osm_list)
}
```

``` r
LandUseData <-
  opq(bbox = bb_box) |>
  add_osm_feature(key = 'landuse') |>
  osmdata_sf() |>
  keep_valid_osm(key = 'landuse')

saveRDS(LandUseData, here("output/cache-data/LandUseData.rds"), compress = "xz")
```

```{r}
#| echo: FALSE

if(!exists("LandUseData")) {
  LandUseData <- readRDS(here("output/cache-data/LandUseData.rds"))
}
```

```{r}
tm_shape(LandUseData$osm_lines) +
  tm_lines()  +
  tm_basemap("CartoDB.Positron")

for(geom_i in seq_along(LandUseData$osm_multilines$geometry)){
  print(plot(LandUseData$osm_multilines$geometry[geom_i]))
}

plot(LandUseData$osm_multilines$geometry[1])
plot(LandUseData$osm_multilines$geometry[2])
plot(LandUseData$osm_multilines$geometry[3])

LandUseData$osm_lines <- st_centroid(LandUseData$osm_lines, of_largest_polygon = TRUE)
LandUseData$osm_multilines <- st_centroid(LandUseData$osm_multilines, of_largest_polygon = TRUE)

sf::sf_use_s2(FALSE)
tm_shape(LandUseData$osm_lines) +
  tm_symbols() +
  tm_basemap("CartoDB.Positron")

tm_shape(LandUseData$osm_multilines) +
  tm_symbols() +
  tm_basemap("CartoDB.Positron")
sf::sf_use_s2(TRUE)

plot(LandUseData$osm_multilines$geometry)
  
project.nyc.taxi::plot_box(bb_box) +
  
  tm_shape(do.call("rbind", LandUseData[4:8])) +
  tm_symbols(fill = "landuse",
             fill.scale = tm_scale_categorical(n.max = 50))+
  tm_polygons(fill = "landuse",
              fill.scale = tm_scale_categorical(n.max = 50))
```

### Adding features

``` r
ZoneCodeRef <-
  here("output/cache-data/ZoneCodesRef.fst") |>
  read_fst(as.data.table = TRUE)

TrainingSampleZones <-
  
  ZoneCodeRef[TrainingSample,
              on = c("LocationID" = "PULocationID"),
              nomatch = NULL
  ][, setnames(.SD, names(ZoneCodeRef), paste0("PU_",names(ZoneCodeRef)))
    
  ][, ZoneCodeRef[.SD,
                  on = c("LocationID" = "DOLocationID"),
                  nomatch = NULL]
  ][, setnames(.SD, names(ZoneCodeRef), paste0("DO_",names(ZoneCodeRef)))]
```

``` r
# timeDate::listHolidays("US") |>
#   (\(x) paste0("`Days to ", x, "` = ", 
#                "difftime(", x,"(year(request_datetime)), request_datetime, units = 'days')",
#                " |> as.integer()") )() |>
#   paste0(collapse = ",\n") |>
#   cat()

TimeRecipe <-
  recipe(~ ., data = TrainingSampleZones) |>
  
  # Daily cycle
  step_harmonic(request_datetime, 
                frequency = 1,
                cycle_size = 3600*24, 
                keep_original_cols = TRUE) |>
  step_rename(request_datetime_sin_daily = request_datetime_sin_1,
              request_datetime_cos_daily = request_datetime_cos_1) |>
  
  # Weekly cycle
  step_harmonic(request_datetime, 
                frequency = 1, 
                cycle_size = 3600*24*7, 
                keep_original_cols = TRUE) %>%
  step_rename(request_datetime_sin_weekly = request_datetime_sin_1,
              request_datetime_cos_weekly = request_datetime_cos_1) |>
  
  # Extracting additional information
  step_date(request_datetime,
            features = c("year",
                         "doy",
                         "week",
                         "decimal",
                         "semester", 
                         "quarter",
                         "dow", 
                         "month")) |>
  step_holiday(request_datetime,
               holidays = c('USChristmasDay',
                            'USColumbusDay',
                            'USCPulaskisBirthday',
                            'USDecorationMemorialDay',
                            'USElectionDay',
                            'USGoodFriday',
                            'USInaugurationDay',
                            'USIndependenceDay',
                            'USJuneteenthNationalIndependenceDay',
                            'USLaborDay',
                            'USLincolnsBirthday',
                            'USMemorialDay',
                            'USMLKingsBirthday',
                            'USNewYearsDay',
                            'USPresidentsDay',
                            'USThanksgivingDay',
                            'USVeteransDay',
                            'USWashingtonsBirthday')) |>
  step_mutate(request_datetime_am = am(request_datetime),
              request_datetime_pm = pm(request_datetime),
              `Days to USChristmasDay` = difftime(USChristmasDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USColumbusDay` = difftime(USColumbusDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USCPulaskisBirthday` = difftime(USCPulaskisBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USDecorationMemorialDay` = difftime(USDecorationMemorialDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USElectionDay` = difftime(USElectionDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USGoodFriday` = difftime(USGoodFriday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USInaugurationDay` = difftime(USInaugurationDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USIndependenceDay` = difftime(USIndependenceDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USJuneteenthNationalIndependenceDay` = difftime(USJuneteenthNationalIndependenceDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USLaborDay` = difftime(USLaborDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USLincolnsBirthday` = difftime(USLincolnsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USMemorialDay` = difftime(USMemorialDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USMLKingsBirthday` = difftime(USMLKingsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USNewYearsDay` = difftime(USNewYearsDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USPresidentsDay` = difftime(USPresidentsDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USThanksgivingDay` = difftime(USThanksgivingDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USVeteransDay` = difftime(USVeteransDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USWashingtonsBirthday` = difftime(USWashingtonsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer())


TrainingSampleZonesTime <-
  prep(TimeRecipe) |>
  bake(new_data = NULL)

glimpse(TrainingSampleZonesTime)
```
