---
title: "Expanding Geospatial Information"
editor_options: 
  chunk_output_type: console
---

## Importing libraries

```{r}
library(here)
library(fst)
library(data.table)
library(sf)
library(tmap)
library(osmdata)

tmap_mode("view")
BoroughColors <- c(
  'Manhattan' = '#e41a1c',
  'Queens' = '#377eb8',
  'Brooklyn'= '#4daf4a'
)
```

## Geocoding the data

```{r}
#| eval: false

# Importing CSV

ZoneCodesRef <- fread(
  here("raw-data/taxi_zone_lookup.csv"),
  colClasses = c("integer",
                 "character",
                 "character",
                 "character")
)


ZoneCodesRefFilter <- ZoneCodesRef[
  c("Queens", "Brooklyn", "Manhattan"),
  on = "Borough",
  nomatch = NULL
]


# Defining the full address to find
# making sure that only use the second part of zone with /
# to get better results when geocoding
ZoneCodesRefFilter[
  j = Address := paste(sub("^.+\\/", "", Zone, TRUE),
                       Borough,
                       "New York",
                       "United States",
                       sep = ", ")
]


ZoneCodesRefSf <-
  
  # Getting Long and Lat
  tidygeocoder::geocode_combine(ZoneCodesRefFilter,
                                queries = list(list(method = 'osm'),
                                               list(method = 'geocodio')),
                                global_params = list(address = 'Address')) |>
  
  # Defining SF as CRS WGS84 (EPSG:4326)
  st_as_sf(coords = c("long", "lat"),
           crs = 4326)


# Saving the result to use later
st_write(ZoneCodesRefSf, here("output/cache-data/ZoneCodesRefSf.geojson"))
```


```{r}
#| echo: false

ZoneCodesRefSf <- st_read(here("output/cache-data/ZoneCodesRefSf.geojson"))
```


```{r}
tm_shape(ZoneCodesRefSf) +
  tm_symbols(size = 0.5,
             fill.scale = tm_scale_categorical(values = BoroughColors),
             fill = "Borough",
             fill_alpha = 0.65) +
  tm_basemap("CartoDB.Positron")
```

6. As we could see, the __*Van Cortlandt Village*__ doesn't seems to be correct as it was placed too far way in the last chart. To solve this let's manually see the correct location using __Google Maps__.

```{r}
ZoneCodesRefSfClean <- ZoneCodesRefSf

st_geometry(ZoneCodesRefSfClean)[which(ZoneCodesRefSf$Zone == "Central Harlem North")] <- st_point(
  c(-73.93871445391756, 40.82304561189441)
)

st_geometry(ZoneCodesRefSfClean)[which(ZoneCodesRefSf$Zone == "Lenox Hill West")] <- st_point(
  c(-73.96538508986032, 40.76921775554697)
)

st_geometry(ZoneCodesRefSfClean)[which(ZoneCodesRefSf$Zone == "Upper West Side South")] <- st_point(
  c(-73.98549143859647, 40.772275233655805)
)

st_geometry(ZoneCodesRefSfClean)[which(ZoneCodesRefSf$Zone == "Upper West Side North")] <- st_point(
  c(-73.9647633273013, 40.80194135268728)
)


tm_shape(ZoneCodesRefSfClean) +
  tm_symbols(size = 0.5,
             fill.scale = tm_scale_categorical(values = BoroughColors),
             fill = "Borough",
             fill_alpha = 0.65) +
  tm_basemap("CartoDB.Positron")
```

7. Saving results as geojson.

```{r}
#| eval: false

st_write(ZoneCodesRefSfClean, here("output/cache-data/ZoneCodesRefSfClean.geojson"))
```


```{r}
# Compute the overall bounding box
bb_box <-
  lapply(X = c("Manhattan, NY", "Brooklyn, NY", "Queens, NY"),
         FUN = osmdata::getbb) |>
  do.call(what = "cbind") |>
  (\(x) cbind(min = apply(x, 1, min),
              max = apply(x, 1, max)))()

project.nyc.taxi::plot_box(bb_box) +
  tmap::tm_shape(ZoneCodesRefSfClean) +
  tmap::tm_symbols(size = 0.5,
                   fill.scale = tm_scale_categorical(values = BoroughColors),
                   fill = "Borough",
                   fill_alpha = 0.65)
```

### Getting data

```{r}
LandUseData <-
  opq(bbox = bb_box) |>
  add_osm_feature(key = 'landuse') |>
  osmdata_sf() |>
  (\(df) rbind(df$osm_polygons["landuse"],
               df$osm_multipolygons["landuse"]))()

# Saving the result to use later
st_write(LandUseData, here("output/cache-data/LandUseData.geojson"), append = FALSE)
```

```{r}
#| echo: false

LandUseData <- st_read(here("output/cache-data/LandUseData.geojson"))
```

```{r}
# Now combined_polygons is an sf object with landuse and geometry columns
tm_shape(LandUseData) +
  tm_polygons(fill = "landuse",
              fill.scale = tm_scale_categorical()) +
  tm_basemap("CartoDB.Positron")
```



### Geospatial features

   *   **From OpenStreetMap (OSM):**
        *   **Count of Points of Interest (POIs)** like bars, restaurants, cafes, hotels, offices, residential buildings.
        *   **OSM Tags:** `amenity=*`, `tourism=*`, `shop=*`, `office=*`, `building=*` etc.
        
        *   **Length of roads** Total road length within a given radius of the pick up and drop off locations, also consider the type of road (`highway=*`)
        *   Density of different types of roads (e.g. highways, main roads, residential streets).
        *   Average speed limit of roads in the area.
         *   **OSM Tags:** `highway=*`, `maxspeed=*`
         
        *   **Number of intersections within a certain radius**
        *   **Distance to Specific Landmarks** like airports, major parks, or business centers.
        
        
        *   **Number and type (subway, bus, train) of transit stops**
        *   Distance to the nearest subway or train station
        *   Number of subway/train stations within a radius
        *   Distance to the closest bus stop or terminal.
    *   **OSM Tags:** `railway=*`, `public_transport=*`, `bus_stop=*`

   *   **From the US Census Bureau (using the Census API):**
        *   **Population Density of the census tract**
        *   **Median household income in the census tract**
        *   **Percentage of Workers Using Public Transportation:**
        *   **proportion of commuters or commuter zone*


4.  **Land Use:**
    *   **Why it matters:** Land use type affects the nature of activity in an area (residential, commercial, industrial).
        *   **Features to extract:**
            *   Percentage of land covered by different categories like parks, commercial areas, industrial areas, residential areas within a radius of the pickup/dropoff locations.
    *   **OSM Tags:** `landuse=*`

5.  **Building Footprints/Density:**
     *  **Why it matters:** The density of buildings often implies how busy a given area is
        *   **Features to extract:**
            *   Count of buildings within a specified radius
            *   Building footprint density (square meters of building area per sq km)
    *    **OSM Tags:** `building=*`

6.  **Elevation/Topography:**
    *   **Why it matters:** Elevation changes could affect taxi trip times or cost for short trips
        *   **Features to extract:**
            *   Average or maximum elevation within a radius of pick up and drop off
            *   Changes in elevation of routes between pick up and drop off
    *   **OSM Data:** you can obtain this information from DEM (Digital Elevation Models), which you can find on raster format.

**How to Use the Features:**

*   **Radius-Based Aggregation:** Most OSM features will be aggregated within a radius around each pick-up and drop-off location. Experiment with different radius sizes (e.g., 0.5 km, 1 km, 2 km) to see which scales capture the most relevant information.
*   **Feature Engineering:** Combine different OSM features. For instance, you might calculate a ratio of restaurants to total commercial buildings.
*   **Join:** Use the `PULocationID` and `DOLocationID` to associate each set of OSM features to each trip from your dataset.

**Why These Features Will Help Your Model:**

*   **Contextual Awareness:** They provide more context about why certain locations are busy or have higher earnings per hour.
*   **Improved Accuracy:** By capturing the built environment and activity around pick-up and drop-off locations, your model will have more fine-grained and predictive features.
*   **Ability to Generalize:** OSM data is widely available and might help your model generalize better to different areas or future data.

**Tools to Extract OSM data:**

*   **Overpass API:** Great for custom queries.
*   **osmnx library (Python):** Makes it easy to download, analyze and process OSM data.

**Important Considerations:**

*   **Data Volume:** OSM data can be very large. You might need to optimize the data extraction and storage process.
*   **Feature Selection:** Be prepared to experiment and select the most relevant features.
*   **Computational Cost:** Processing spatial data can be computationally expensive, so make sure your system is ready to process all this data.

By incorporating these OSM features, you can add a layer of contextual richness to your data, which will significantly improve your model's ability to predict trip performance and provide valuable insights to taxi drivers.




### Adding features

```r
ZoneCodeRef <-
  here("output/cache-data/ZoneCodesRef.fst") |>
  read_fst(as.data.table = TRUE)

TrainingSampleZones <-
  
  ZoneCodeRef[TrainingSample,
              on = c("LocationID" = "PULocationID"),
              nomatch = NULL
  ][, setnames(.SD, names(ZoneCodeRef), paste0("PU_",names(ZoneCodeRef)))
    
  ][, ZoneCodeRef[.SD,
                  on = c("LocationID" = "DOLocationID"),
                  nomatch = NULL]
  ][, setnames(.SD, names(ZoneCodeRef), paste0("DO_",names(ZoneCodeRef)))]
```


```r
# timeDate::listHolidays("US") |>
#   (\(x) paste0("`Days to ", x, "` = ", 
#                "difftime(", x,"(year(request_datetime)), request_datetime, units = 'days')",
#                " |> as.integer()") )() |>
#   paste0(collapse = ",\n") |>
#   cat()

TimeRecipe <-
  recipe(~ ., data = TrainingSampleZones) |>
  
  # Daily cycle
  step_harmonic(request_datetime, 
                frequency = 1,
                cycle_size = 3600*24, 
                keep_original_cols = TRUE) |>
  step_rename(request_datetime_sin_daily = request_datetime_sin_1,
              request_datetime_cos_daily = request_datetime_cos_1) |>
  
  # Weekly cycle
  step_harmonic(request_datetime, 
                frequency = 1, 
                cycle_size = 3600*24*7, 
                keep_original_cols = TRUE) %>%
  step_rename(request_datetime_sin_weekly = request_datetime_sin_1,
              request_datetime_cos_weekly = request_datetime_cos_1) |>
  
  # Extracting additional information
  step_date(request_datetime,
            features = c("year",
                         "doy",
                         "week",
                         "decimal",
                         "semester", 
                         "quarter",
                         "dow", 
                         "month")) |>
  step_holiday(request_datetime,
               holidays = c('USChristmasDay',
                            'USColumbusDay',
                            'USCPulaskisBirthday',
                            'USDecorationMemorialDay',
                            'USElectionDay',
                            'USGoodFriday',
                            'USInaugurationDay',
                            'USIndependenceDay',
                            'USJuneteenthNationalIndependenceDay',
                            'USLaborDay',
                            'USLincolnsBirthday',
                            'USMemorialDay',
                            'USMLKingsBirthday',
                            'USNewYearsDay',
                            'USPresidentsDay',
                            'USThanksgivingDay',
                            'USVeteransDay',
                            'USWashingtonsBirthday')) |>
  step_mutate(request_datetime_am = am(request_datetime),
              request_datetime_pm = pm(request_datetime),
              `Days to USChristmasDay` = difftime(USChristmasDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USColumbusDay` = difftime(USColumbusDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USCPulaskisBirthday` = difftime(USCPulaskisBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USDecorationMemorialDay` = difftime(USDecorationMemorialDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USElectionDay` = difftime(USElectionDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USGoodFriday` = difftime(USGoodFriday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USInaugurationDay` = difftime(USInaugurationDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USIndependenceDay` = difftime(USIndependenceDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USJuneteenthNationalIndependenceDay` = difftime(USJuneteenthNationalIndependenceDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USLaborDay` = difftime(USLaborDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USLincolnsBirthday` = difftime(USLincolnsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USMemorialDay` = difftime(USMemorialDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USMLKingsBirthday` = difftime(USMLKingsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USNewYearsDay` = difftime(USNewYearsDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USPresidentsDay` = difftime(USPresidentsDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USThanksgivingDay` = difftime(USThanksgivingDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USVeteransDay` = difftime(USVeteransDay(year(request_datetime)), request_datetime, units = 'days') |> as.integer(),
              `Days to USWashingtonsBirthday` = difftime(USWashingtonsBirthday(year(request_datetime)), request_datetime, units = 'days') |> as.integer())


TrainingSampleZonesTime <-
  prep(TimeRecipe) |>
  bake(new_data = NULL)

glimpse(TrainingSampleZonesTime)
```


