---
title: "From Predictions to Policies: Integrating ML into Stochastic Optimization"
subtitle: "Evaluating the ADP Policy in Full-Day Monte-Carlo Simulation"
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This chapter closes the Approximate Dynamic Programming (ADP) loop. We convert the probabilistic XGBoost classifier into a **tunable threshold policy** $\pi^\tau$ and re-evaluate it inside the *exact same* full-day empirical Monte-Carlo simulator used for the baseline in Chapter 4. The result is a directly comparable hourly-wage lift under the identical data-driven uncertainty model $\mathcal{C}_t^n$.

## Updated Mathematical Model

The state, exogenous information, and transition functions remain identical to Chapter 4. Only the **decision rule** changes.

**Decision variable**  
$$
x_t = \mathbb{I} \{ \hat{p}_t(S_t) \ge \tau \}
$$

where $\hat{p}_t(S_t)$ is the XGBoost probability that the *current offered trip* (sampled from $\mathcal{C}_t^n$) belongs to the high-value class, and $\tau \in [0,1]$ is the tunable policy parameter.


**Policy definition**  
$$
\pi^\tau(S_t) = 
\begin{cases}
1 & \text{if } \hat{p}_t(S_t) \ge \tau \quad (\text{accept current trip}) \\
0 & \text{otherwise} \quad (\text{reject and expand search limits})
\end{cases}
$$

**Objective function (to be maximized)**  

We seek to **maximize** the expected hourly wage under the data-driven uncertainty $\mathcal{C}_t^n$:

$$
F^{\pi^\tau}(S_0^n) = \mathbb{E}\left[ 
\frac{\sum_{t=1}^{T^n} (\text{Trip\_Earning}_{t+1}^n \cdot x_t)}
{(\text{RequestTime}_{T^n}^n + \text{trip\_time}_{T^n}^n) - \text{EndTime}_{I_0^n}} 
\;\middle|\; S_0^n \right]
$$

The expectation is estimated via Monte-Carlo simulation using **five independent random seeds** for each of the 1,061 starting states $S_0^n$. This provides both a reproducible point estimate and a confidence interval that is directly comparable to the baseline.

**Important operational detail** 

When $x_t = 0$ the driver does **not** advance in position or time; the search limits expand exactly as in the baseline rejection case (`Trip_Time_Limit += 1 or 2 min`, `Trip_Dist_Limit += 2 mi`). This faithfully reproduces the 15-minute lookahead horizon that generated the training labels and keeps the simulation apples-to-apples with Chapter 4.

## Loading the Production XGBoost Workflow

```{r}
## To manage relative paths
library(here)
library(pins)
library(qs2)

## Manipulate data
library(data.table)
library(DBI)
library(duckdb)
library(lubridate)
library(timeDate)

## To run predictions
library(tidymodels)

## To show results
library(ggtext)
library(scales)

## Custom functions & board
devtools::load_all()
BoardLocal <- board_folder(here("../NycTaxiPins/Board"))

# Loading params
Params <- yaml::read_yaml(here("params.yml"))

# Sampled data (Chapter 4)
SimulationStartDay <- pin_read(
  BoardLocal,
  name = "SimulationStartDay"
)
SimulationHourlyWage <- pin_read(
  BoardLocal,
  name = "SimulationHourlyWage"
)

BaseLineHourlyWage <- pin_read(BoardLocal, "BaseLineHourlyWage")

# Production workflow (Chapter 8)
XgboostFinalWorkflowTailorFitted <- pin_read(
  BoardLocal,
  name = "XgboostFinalWorkflowTailorFitted"
)

InitalBestThresholdValue <- pin_read(BoardLocal, "BestThresholdValue")
```

## Threshold Optimization under Stochastic Uncertainty

### Why simulation-based tuning is essential

The validation threshold ($\approx `r round(InitalBestThresholdValue, 2)`$) was optimal on held-out *historical* trips under a static cost matrix. The true economic optimum must be found under the *dynamic* uncertainty $\mathcal{C}_t^n$ of the full-day simulator. We therefore perform a grid search *inside* the Monte-Carlo engine itself.


### Seeds strategy (reproducibility + fairness)

- **Tuning phase**: Five new seeds (`new_seeds`) to avoid overfitting to the exact randomness of the baseline.  
- **Final evaluation**: The best $\tau^*$ is re-run with the original baseline seeds `c(8199, 50, 7450, 1430, 9370)` so the confidence interval is directly comparable to Chapter 4.

```{r}
original_seeds <- c(8199, 50, 7450, 1430, 9370) # exactly as Chapter 4

set.seed(987654) # new seed for tuning reproducibility
(new_seeds <- sample.int(1e6, 5)) # 5 new seeds for grid search
```


### Computational cost and adaptive sampling strategy

The grid search across the threshold space $\tau \in [0.05, 0.95]$ required approximately **10 days** of continuous execution. To balance rigor with practicality we used all five seeds for $\tau < 0.65$ and the first three seeds for $\tau \ge 0.65$ once convergence patterns became stable.

To balance statistical rigor with practical time constraints, an adaptive sampling strategy was employed to optimize the simulation throughput:

* **For $\tau < 0.65$:** All **5 seeds** were utilized to ensure maximum precision in the region where the policy's performance was most dynamic.
* **For $\tau \ge 0.65$:** After observing stable convergence patterns and a clear performance trend, the execution was streamlined to use only the **first three seeds**. 

This pragmatic adjustment allowed the sweep to reach the upper bound of $0.95$ without compromising the identification of the global optimum, ensuring that the most "expensive" compute time was focused where the variance mattered most.


### Grid-search code

The simulator is called with `fitted_wf = XgboostFinalWorkflowTailorFitted` and `threshold = \tau`. The process took approximately **10 days** on the available hardware.

```{r}
#| eval: false
#| label: tau-grid-search

TauGrid <- seq(0.05, 0.95, by = 0.05)

con <- dbConnect(duckdb(), dbdir = here("../NycTaxiBigFiles/my-db.duckdb"))

for (i in seq_along(TauGrid)) {
  # Running simulation
  SimResults <- simulate_trips(
    con,
    SimulationStartDay, # same 1,061 starting trips as baseline
    seeds = new_seeds, # NEW seeds for tuning
    fitted_wf = XgboostFinalWorkflowTailorFitted,
    threshold = TauGrid[i]
  )

  # Adding initial_day_time
  SimResults[
    SimulationStartDay,
    on = c("simulation_id" = "trip_id"),
    initial_day_time := request_datetime + lubridate::seconds(trip_time)
  ]

  SimResultsSummary <- SimResults[,
    .(
      n_trips = .N,
      total_hours_worked = difftime(
        max(sim_dropoff_datetime),
        min(initial_day_time),
        unit = "hours"
      ) |>
        as.double(),
      total_earnings = sum(sim_driver_pay, na.rm = TRUE) +
        sum(sim_tips, na.rm = TRUE)
    ),
    by = c("simulation_id", "simulation_seed")
  ][, daily_hourly_wage := total_earnings / total_hours_worked][,
    j = c(
      setNames(lapply(.SD, mean), paste0(names(.SD), "_mean")),
      setNames(lapply(.SD, sd), paste0(names(.SD), "_sd"))
    ),
    .SDcols = !c("simulation_seed"),
    by = "simulation_id"
  ]

  setcolorder(
    SimResultsSummary,
    c(
      "simulation_id",
      paste0("n_trips", c("_mean", "_sd")),
      paste0("total_hours_worked", c("_mean", "_sd")),
      paste0("total_earnings", c("_mean", "_sd")),
      paste0("daily_hourly_wage", c("_mean", "_sd"))
    )
  )

  pin_write(
    BoardLocal,
    SimResultsSummary,
    paste0("SimResultsSummary_", TauGrid[i]),
    type = "qs2"
  )
}

dbDisconnect(con, shutdown = TRUE)
```

::: {.callout-note}
The bash output below validates the execution timeline, showing the multi-day progression from the first logs on March 22nd to the final thresholds on April 3rd.
:::

```bash
ls -lhad ../NycTaxiPins/Board/SimResultsSummary_0.*

angelfeliz@debian:~/r-projects/NycTaxi$ ls -lhad ../NycTaxiPins/Board/SimResultsSummary_0.*
drwxr-xr-x 3 root root 4.0K Mar 22 11:05 ../NycTaxiPins/Board/SimResultsSummary_0.05
drwxr-xr-x 3 root root 4.0K Mar 22 19:06 ../NycTaxiPins/Board/SimResultsSummary_0.1
drwxr-xr-x 3 root root 4.0K Mar 23 04:04 ../NycTaxiPins/Board/SimResultsSummary_0.15
drwxr-xr-x 3 root root 4.0K Mar 23 12:22 ../NycTaxiPins/Board/SimResultsSummary_0.2
drwxr-xr-x 3 root root 4.0K Mar 23 21:20 ../NycTaxiPins/Board/SimResultsSummary_0.25
drwxr-xr-x 3 root root 4.0K Mar 24 06:31 ../NycTaxiPins/Board/SimResultsSummary_0.3
drwxr-xr-x 3 root root 4.0K Mar 24 16:15 ../NycTaxiPins/Board/SimResultsSummary_0.35
drwxr-xr-x 3 root root 4.0K Mar 25 02:39 ../NycTaxiPins/Board/SimResultsSummary_0.4
drwxr-xr-x 3 root root 4.0K Mar 25 13:46 ../NycTaxiPins/Board/SimResultsSummary_0.45
drwxr-xr-x 3 root root 4.0K Mar 26 01:28 ../NycTaxiPins/Board/SimResultsSummary_0.5
drwxr-xr-x 3 root root 4.0K Mar 26 14:03 ../NycTaxiPins/Board/SimResultsSummary_0.55
drwxr-xr-x 3 root root 4.0K Mar 27 03:36 ../NycTaxiPins/Board/SimResultsSummary_0.6
drwxr-xr-x 3 root root 4.0K Mar 30 12:11 ../NycTaxiPins/Board/SimResultsSummary_0.65
drwxr-xr-x 3 root root 4.0K Mar 30 21:51 ../NycTaxiPins/Board/SimResultsSummary_0.7
drwxr-xr-x 3 root root 4.0K Mar 31 08:51 ../NycTaxiPins/Board/SimResultsSummary_0.75
drwxr-xr-x 3 root root 4.0K Mar 31 21:35 ../NycTaxiPins/Board/SimResultsSummary_0.8
drwxr-xr-x 3 root root 4.0K Apr  1 12:06 ../NycTaxiPins/Board/SimResultsSummary_0.85
drwxr-xr-x 3 root root 4.0K Apr  2 07:15 ../NycTaxiPins/Board/SimResultsSummary_0.9
drwxr-xr-x 3 root root 4.0K Apr  3 12:21 ../NycTaxiPins/Board/SimResultsSummary_0.95
```

### Exploring the Results

```{r}
SimResultsSummary <-
  pin_list(BoardLocal) |>
  grep(pattern = "SimResultsSummary_", value = TRUE) |>
  lapply(
    FUN = \(x) {
      pin_read(BoardLocal, x)[,
        tau_used := sub("SimResultsSummary_", "", x) |> as.double()
      ]
    }
  ) |>
  rbindlist(use.names = TRUE)

HourlyWageByTau <-
  SimResultsSummary[,
    .(
      mean_n_trips = mean(n_trips_mean),
      se_n_trips = sd(n_trips_mean) / sqrt(.N),
      mean_wage = mean(daily_hourly_wage_mean),
      se_wage = sd(daily_hourly_wage_mean) / sqrt(.N)
    ),
    by = "tau_used"
  ]

BestResult <- HourlyWageByTau[which.max(mean_wage), ]
```

The results are striking. Contrary to the common intuition that drivers should “accept most trips and reject only the worst,” hourly wage **increases monotonically** with the threshold $\tau$. It rises from \$`r round(HourlyWageByTau[1L, mean_wage], 2)` at $\tau=0.05$ to \$`r round(BestResult$mean_wage, 2)` at $\tau= 0.9$ — a **`r round((BestResult$mean_wage - BaseLineHourlyWage) / BaseLineHourlyWage * 100, 2)`% lift** over the baseline of \$`r round(BaseLineHourlyWage, 2)`.

This counter-intuitive outcome is a powerful validation of the entire ADP pipeline. The lookahead-based labeling process (Chapter 5) created high-quality supervised targets that approximate the optimal short-horizon decision $x_t^*$. The resulting PFA has learned subtle, high-value trip characteristics that the myopic baseline misses. By becoming more selective, the driver trades a modest increase in waiting time—already captured by the expanding-search logic—for substantially higher earnings per hour worked.

```{r}
#| label: Simulated Hourly Wage vs. Threshold
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

ggplot(HourlyWageByTau, aes(tau_used, mean_wage)) +
  geom_vline(xintercept = BestResult$tau_used, linetype = 2) +
  geom_line(
    color = Params$ColorHighlight,
    linewidth = 1
  ) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = mean_wage - 1.96 * se_wage, ymax = mean_wage + 1.96 * se_wage),
    width = 0.015,
    alpha = 0.6
  ) +
  scale_x_continuous(breaks = breaks_width(0.1)) +
  scale_y_continuous(breaks = breaks_width(1), label = dollar_format(1)) +
  labs(
    title = "Simulated Hourly Wage vs. Threshold",
    subtitle = "Peak at optimal operational threshold",
    x = "Threshold",
    y = "Mean Hourly Wage"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

By exploring the average number of trips taken we see that, starting at $\tau=0.7$, the number of trips *increases*. This indicates that the policy is successfully avoiding long, low-value trips, thereby freeing time for more short, high-value trips within the same working hours.

```{r}
#| label: Simulated Number of Trips vs. Threshold
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

ggplot(HourlyWageByTau, aes(tau_used, mean_n_trips)) +
  geom_vline(xintercept = BestResult$tau_used, linetype = 2) +
  geom_line(color = Params$ColorHighlight, linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_n_trips - 1.96 * se_n_trips,
      ymax = mean_n_trips + 1.96 * se_n_trips
    ),
    width = 0.015,
    alpha = 0.6
  ) +
  scale_x_continuous(breaks = breaks_width(0.1)) +
  scale_y_continuous(breaks = breaks_width(0.5), label = dollar_format(1)) +
  labs(
    title = "Simulated Number of Trips vs. Threshold",
    subtitle = "Finding short trips is critical to increase earnings",
    x = "Threshold",
    y = "Mean Hourly Wage"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

::: {.callout-warning}
The simulation assumes abundant trip supply (“more passengers than taxis”). In reality, widespread selectivity could increase city-wide waiting times. Modeling supply-side dynamics remains an important direction for future work.
:::


## Final Policy Evaluation

We now re-run the simulator **only** at the optimal $\tau^*$ using the **exact same seeds** as the baseline.

```{r}
#| eval: false

con <- dbConnect(duckdb(), dbdir = here("../NycTaxiBigFiles/my-db.duckdb"))

FinalPfaResults <- simulate_trips(
  con,
  SimulationStartDay,
  seeds = original_seeds,
  fitted_wf = XgboostFinalWorkflowTailorFitted,
  threshold = BestResult$tau_used
)

dbDisconnect(con, shutdown = TRUE)

pin_write(BoardLocal, FinalPfaResults, "FinalPfaResults", type = "qs2")
```

::: {.callout-note}
This final run took approximately **2 days** on the available hardware.
:::

### Computing the new hourly wage & CI

An important aspect visible in the new policy is its larger interquartile range. We expect this variability will be reduced once we add a strategic policy that selects the best starting zone and hour. We also anticipate that the second policy will help us reach the full **20 % lift** targeted by the project.

The boxplot below confirms that the new mean hourly wage remained stable around the value obtained during grid search, demonstrating that the policy is robust and ready for production.

```{r}
FinalPfaResults <- pin_read(BoardLocal, "FinalPfaResults")

PolicyLevels <- c("Accept All Trips", "Accept Based On Policy")

FinalPfaResultsSummary <- rbind(
  sim_start_trip_summary(SimulationHourlyWage, SimulationStartDay)[,
    strategy := PolicyLevels[1L]
  ],
  sim_start_trip_summary(FinalPfaResults, SimulationStartDay)[,
    strategy := PolicyLevels[2L]
  ],
  fill = FALSE
)

FinalPfaResultsSummary[, strategy := factor(strategy, PolicyLevels)]
```

```{r}
#| label: Mean Hourly Wage after policy
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

PolicyLevelsFillColors <- c(Params$ColorGray, Params$ColorHighlight)
setattr(PolicyLevelsFillColors, "names", PolicyLevels)

FinalPfaMeans <-
  FinalPfaResultsSummary[,
    .(daily_hourly_wage_mean = mean(daily_hourly_wage_mean)),
    keyby = "strategy"
  ][, setattr(daily_hourly_wage_mean, "names", as.character(strategy))]

ggplot(FinalPfaResultsSummary, aes(strategy, daily_hourly_wage_mean)) +
  geom_boxplot(aes(fill = strategy), width = 0.5) +
  geom_hline(yintercept = FinalPfaMeans[2L], linetype = 2) +
  scale_fill_manual(
    values = PolicyLevelsFillColors
  ) +
  scale_y_continuous(breaks = breaks_width(5), label = dollar_format(1)) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**",
      PolicyLevels[2L],
      "**</span> ",
      "**is the best policy**"
    ),
    subtitle = paste0(
      "The new mean kept around **",
      dollar(BestResult$mean_wage, accuracy = 1),
      "** confirming stable results"
    ),
    x = "Strategy Implemented",
    y = "Mean Hourly Wage"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(size = 14, family = Params$BaseFontFamily),
    plot.subtitle = element_markdown(size = 12, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```

When we examine the improvement on a per-simulation-day basis, the gains are remarkably consistent: most simulated days earn **at least $2 extra per hour**, and the distribution of improvements is approximately normal.

```{r}
MeanPolicyDifference <- FinalPfaResultsSummary[,
  .(
    mean_difference = fifelse(
      strategy == PolicyLevels[1L],
      -daily_hourly_wage_mean,
      daily_hourly_wage_mean
    ) |>
      sum()
  ),
  by = "simulation_id"
]

shapiro.test(MeanPolicyDifference$mean_difference)
```

```{r}
#| label: Prop of improved simulation days
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5

InterceptToValidate <- 2
PropSimoOverIntercept <- MeanPolicyDifference[, mean(
  mean_difference >= InterceptToValidate
)]

ggplot(MeanPolicyDifference) +
  geom_histogram(aes(mean_difference), bins = 35) +
  geom_vline(xintercept = InterceptToValidate, linetype = 2) +
  scale_x_continuous(n.breaks = 10, label = dollar_format(1)) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**",
      percent(PropSimoOverIntercept, accuracy = 1),
      "**</span> ",
      "**of simulated days are earning at least 2 dollars extra**"
    ),
    x = "Mean Hourly Wage Increase"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold")
  )
```


# Key Takeaways

This chapter completes the ADP loop for the trip-acceptance policy. The combination of lookahead labeling, a well-calibrated PFA, and simulation-based threshold optimization has produced a practically usable, profit-driven decision rule that demonstrably outperforms the myopic baseline under the same uncertainty model.

```{r}
#| echo: false
#| fig.width: 9
#| fig.height: 1
#| out.width: "100%"

DiagrammeR::grViz(
  "
digraph ADP_Pipeline {
  graph [layout = dot, rankdir = LR, fontsize = 12]
  node [shape = box, style = filled, fontname = Arial]

  Baseline [label = 'Baseline', fillcolor = '#D3CDF7']
  Labeled [label = 'Labeled Data', fillcolor = '#D3CDF7']
  TrainedModel [label = 'Trained PFA', fillcolor = '#D3CDF7']
  SimPolicy [label = 'Simulation-Optimized Policy', fillcolor = '#D3CDF7']
  Quantified [label = 'Quantified Improvement', fillcolor = '#D3CDF7']

  Baseline -> Labeled
  Labeled -> TrainedModel
  TrainedModel -> SimPolicy
  SimPolicy -> Quantified
}
"
)
```



