---
title: "Maximizing Driver Earnings by Selecting the Best Configuration and Time to Start"
subtitle: "A Simulation-Driven Optimization of Ride-Hail Policies Using Tree-Based Interactions"
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once have a policy to decide if a trip should be taken or rejected then it's important to validate if any of the next factors could cause an increase of decrease on taxi results after blocking the **company** for taxi driver:

- Start zone
- Days of the week
- Hour of the day

## Defining combinations to simulate

As we have 196 zones to evaluete, 3 days to evaluete for each day of the week (7), 2 taxi companies and we divide the day in 8 moments to measure (one each 3 hours), we end with an experiment of **65,856 days** to simulate.

```{r}
#| label: Defining combinations
# Loading functions ----

## To manage relative paths
library(here)
library(pins)
library(qs2)

## Manipulate data
library(data.table)
library(sf)
library(DBI)
library(duckdb)
library(lubridate)
library(timeDate)
library(ggtext)
library(scales)
library(gt)

## To run predictions
library(tidymodels)
library(embed)
library(themis)

## Custom functions, board & params
devtools::load_all()
BoardLocal <- board_folder(here("../NycTaxiPins/Board"))
Params <- yaml::read_yaml(here("params.yml"))


# Defining combinations ----

## Getting shapes to use
ZonesShapes <-
  pin_read(BoardLocal, "ZonesShapes") |>
  (\(df) {
    df[
      df[["borough"]] %chin% names(Params$BoroughColors),
    ]
  })() |>
  st_cast(to = "MULTIPOLYGON")

setDT(ZonesShapes)
ZonesShapes[, LocationID := as.character(LocationID)]

UniqueZoneIds <- unique(ZonesShapes$LocationID) |> sort()


## Defining days to sample
YearDays <- data.table(
  days = seq.Date(make_date(2023), make_date(2023, 12, 31), by = "day")
)
YearDays[, wday := lubridate::wday(days)]
set.seed(4879)
SampledDays <- YearDays[, .SD[sample.int(.N, 3L)], by = "wday"]


## Pre-build the big design matrix once
DataToSimulate <- CJ(
  DOLocationID = UniqueZoneIds,
  hvfhs_license_num = c("HV0003", "HV0005"),
  calendar_day = as.character(SampledDays$days),
  time_hour = seq(0, 21, length.out = 8)
)
DataToSimulate[, `:=`(
  trip_id = as.numeric(1:.N),
  request_datetime = paste(calendar_day, time_hour) |> ymd_h(),
  trip_time = 0
)]

head(DataToSimulate, 8)
```

## Running simulation

As I knew that running the hole process would take me several days I decided to store the results of simulations on disk for each zone. In this way I was able to keep the experiment as I was expecting and stop the process between days to allow the computer to rest between days and redice the risk of loossing all the progress I had problems with memory while using the computer for dealy activities.

To avoid the overhead of using a IDE the next code was source from terminal after connecting with `ssh NycTaxi` to consume the fewer results as possible.

::: {.callout-note}
In the below code you see an assigning the starting zone to `DOLocationID` as the simulation uses the last zone of the configuration trip (the one we are creting) to starting picking trips from that zone.
:::

```{r}
#| eval: false
#| label: Simulation code
## Loading env

# Getting prior policy
AcceptRejectPolicyFitted <-
  pin_read(BoardLocal, "AcceptRejectPolicyFitted")

# Competing renaming columns
DataToSimulate[, `:=`(
  calendar_day = NULL,
  time_hour = NULL,
  PULocationID = "1",
  wav_match_flag = "N",
  driver_pay = 0,
  tips = 0,
  dispatching_base_num = "",
  originating_base_num = "",
  on_scene_datetime = NA_POSIXct_,
  pickup_datetime = NA_POSIXct_,
  dropoff_datetime = NA_POSIXct_,
  trip_miles = 0,
  base_passenger_fare = 0,
  tolls = 0,
  bcf = 0,
  sales_tax = 0,
  congestion_surcharge = 0,
  airport_fee = 0,
  shared_request_flag = "",
  shared_match_flag = "",
  access_a_ride_flag = "",
  wav_request_flag = "",
  month = 1,
  year = 2023,
  performance_per_hour = 0
)]

# Check which zones are already done
AlreadyDone <- pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  gsub(pattern = "InitialZoneSim_ZoneId_", replacement = "")

zones_to_run <- setdiff(UniqueZoneIds, AlreadyDone)

cat(
  "→",
  length(zones_to_run),
  "zones left to simulate (out of",
  length(UniqueZoneIds),
  "total).\n"
)


# === MAIN RESUMABLE LOOP ===
for (zone_i in zones_to_run) {
  cat("   Simulating zone", zone_i, "... ")

  # Filter only this zone
  ZoneData <- DataToSimulate[list(zone_i), on = "DOLocationID"]

  # Fresh connection for each zone
  con <- dbConnect(duckdb(), dbdir = here("../NycTaxiBigFiles/my-db.duckdb"))

  ZoneSim <- simulate_trips(
    con,
    ZoneData,
    seeds = 4878,
    fitted_wf = AcceptRejectPolicyFitted
  )

  dbDisconnect(con, shutdown = TRUE)

  # Save immediately with clear name
  pin_write(
    BoardLocal,
    ZoneSim,
    paste0("InitialZoneSim_ZoneId_", zone_i),
    type = "qs2"
  )

  cat("DONE (", nrow(ZoneSim), "rows)\n")
}

cat("✅ All zones simulated!\n")
```

Here is the list of zone simulated and we can see that it took around **102 minutes** to complete each zone which mean that I invest around **14 days** of computation to see the next results.

```bash
angelfeliz@debian:~/r-projects/NycTaxi$ ls -ld ../NycTaxiPins/Board/InitialZoneSim_ZoneId_* | wc -l
196
```

Now we just need to summary the results to start the exploration.

```{r}
#| eval: false
#| label: Consolidating results
consolidate_all_simulations <- function(zone_id_data, board, sim_start_days) {
  consolidated_data =
    sim_start_trip_summary(
      pin_read(board, zone_id_data),
      sim_start_days
    )[,
      sim_start_days[, list(
        simulation_id = trip_id,
        taxi_company = fifelse(hvfhs_license_num == 'HV0003', "Uber", "Lyft"),
        start_zone = DOLocationID,
        week_day = wday(calendar_day),
        time_hour
      )][.SD, on = "simulation_id"],
      .SDcols = !patterns("_sd")
    ]

  return(consolidated_data)
}

InitialZoneSimResults <-
  pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  lapply(
    FUN = consolidate_all_simulations,
    board = BoardLocal,
    sim_start_days = DataToSimulate
  ) |>
  rbindlist()

# Adding region info
InitialZoneSimResults[
  ZonesShapes,
  on = c("start_zone" = "LocationID"),
  `:=`(borough = borough, start_zone_name = zone)
]

# Save immediately with clear name
pin_write(
  BoardLocal,
  InitialZoneSimResults,
  "InitialZoneSimResults",
  type = "qs2"
)
```


## Exploring results

Before anything, it's important to confirm if after changing the initial conditions of trips had an effect over the target variable. Just imagene that the we see the same value for all combinations we could stop the question just by making the first chart.

But you can see floowing chart preset a lot variation, it seems like the data is following a **log-normal** due it's right tail. This are really promessing results, the goal behind this experiment is to indentify the starting configurations that can beat the current baseline of **`r here::here("../NycTaxiPins/Board") |> pins::board_folder() |> pins::pin_read( "BaseLineHourlyWage") |> dollar()`** and it's clear based on the chart that the **top 10%** presented higher daily hourly have than the base line.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing distribution
InitialZoneSimResults <- pin_read(
  BoardLocal,
  "InitialZoneSimResults"
)

FinalPfaMeans <- pin_read(
  BoardLocal,
  "FinalPfaMeans"
)


TopDailyHourlyWage <- quantile(
  InitialZoneSimResults$daily_hourly_wage_mean,
  0.90
)

InitialZoneSimResults[,
  is_top_trip := fifelse(
    daily_hourly_wage_mean >= TopDailyHourlyWage,
    "yes",
    "no"
  ) |>
    factor(levels = c("no", "yes"))
]


BreaksToPlot <- c(seq(20, 200, 10), unname(FinalPfaMeans)) |> setdiff(60L)

highlight_text <- function(x) {
  paste0(
    " <span style='color:",
    Params$ColorHighlight,
    ";'>**",
    x,
    "**</span> "
  )
}

ggplot(
  InitialZoneSimResults,
  aes(daily_hourly_wage_mean, fill = is_top_trip)
) +
  geom_histogram(color = "black", bins = 30) +
  geom_vline(
    xintercept = FinalPfaMeans[2L],
    linetype = 2
  ) +
  scale_y_continuous(
    label = label_comma(accuracy = 1)
  ) +
  scale_x_continuous(
    label = label_currency(accuracy = 1),
    breaks = setdiff(BreaksToPlot, FinalPfaMeans[1L])
  ) +
  scale_fill_manual(
    values = c("yes" = Params$ColorHighlight, "no" = Params$ColorGray)
  ) +
  labs(
    title = paste0(
      "**Top**",
      highlight_text("10%"),
      "**of combinations make at least**",
      highlight_text(paste0("_", dollar(TopDailyHourlyWage), " per hour_"))
    ),
    y = "Count",
    x = "Daily Hourly Wage"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily, base_size = 12) +
  theme(
    plot.title = element_markdown(size = 12, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```

As we can see the improment behind the the policy only benefit a **Uber** drivers as the median Hourly Wage for **Lyft** drivers match this the exact value of the baseline, as result our policy will always recomend **Uber** as taxi company.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing taxi company effect
#| fig-height: 6
TaxiCompanyFillColors <- c(
  "Uber" = Params$ColorHighlight,
  "Lyft" = Params$ColorGray
)

ggplot(
  InitialZoneSimResults,
  aes(
    factor(taxi_company, levels = names(TaxiCompanyFillColors)),
    daily_hourly_wage_mean,
    fill = taxi_company
  )
) +
  geom_boxplot(width = 0.5) +
  geom_hline(
    yintercept = FinalPfaMeans[2L],
    linetype = 2
  ) +
  geom_hline(
    yintercept = FinalPfaMeans[1L],
    linetype = 2
  ) +
  scale_fill_manual(
    values = TaxiCompanyFillColors
  ) +
  scale_y_continuous(
    breaks = BreaksToPlot,
    label = dollar_format(1)
  ) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**_Uber_**</span> ",
      "**drivers makes more money than** ",
      "<span style='color:",
      "#5b5b5b",
      ";'>**_Lyft_**</span> **drivers**"
    ),
    x = "Taxi Company",
    y = "Daily Hourly Wage"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(size = 14, family = Params$BaseFontFamily),
    plot.subtitle = element_markdown(size = 12, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```

After isoliating the effect of the taxi company in the hourly wage we can see by selected the zones with highest and lowers median that change in the target variable explanied by starting zone is low due that that we are evaluating **`{r} uniqueN(InitialZoneSimResults$start_zone)` zones** that increase in the median of Uber drivers was only **$2** over or baseline ($`{r} round(FinalPfaMeans[2L], 2)`) after implementing the trip selection model and based on the boxplot we can conclude that zones don't provide enough information to explain the variance present in the target variable.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing zone effect
#| fig-height:  8
TopBottomZonesMedian <- InitialZoneSimResults[,
  list(daily_hourly_wage_mean = median(daily_hourly_wage_mean)),
  by = c("taxi_company", "start_zone")
][order(daily_hourly_wage_mean)][,
  .SD[c(1:10, .N - 0:9)],
  by = "taxi_company",
  .SDcols = "start_zone"
][, InitialZoneSimResults[.SD, on = c("taxi_company", "start_zone")]][,
  start_zone_name := reorder(
    paste0(start_zone_name, "-", taxi_company),
    daily_hourly_wage_mean,
    median
  )
]

ggplot(
  TopBottomZonesMedian,
  aes(daily_hourly_wage_mean, start_zone_name, fill = borough)
) +
  geom_boxplot() +
  geom_vline(xintercept = FinalPfaMeans[1L], linetype = 2) +
  geom_vline(xintercept = FinalPfaMeans[2L], linetype = 2) +
  scale_fill_manual(values = Params$BoroughColors) +
  scale_x_continuous(
    breaks = BreaksToPlot,
    label = dollar_format(1)
  ) +
  scale_y_discrete(label = \(x) sub("-(.|\\s)+$", "", x)) +
  facet_wrap(vars(taxi_company), ncol = 1, scales = "free_y") +
  labs(
    title = "Some difference can be explained by start zone",
    subtitle = "After splitting by taxi company",
    y = "Start Simulation Zone",
    x = "Daily Hourly Wage",
    fill = "Borough"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(
      size = 14,
      face = "bold",
      family = Params$BaseFontFamily
    ),
    plot.subtitle = element_text(
      size = 12,
      face = "italic",
      family = Params$BaseFontFamily
    ),
    axis.text.y = element_text(
      size = 10,
      face = "italic",
      family = Params$BaseFontFamily
    ),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold", family = Params$BaseFontFamily),
    strip.text = element_text(
      size = 12,
      face = "bold",
      family = Params$BaseFontFamily
    )
  )
```

Based on the next chart we can see a strong effect of start hour in the target variable and some moderate interactions with the day of the week as the hourly wage is higher at 0 hours on Sundays and Saturdays, but those increases are not present on week days.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing hour cycles effect
Weekdays <- c(
  "Sunday",
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday"
)

BetterThanModelFillColors <- c(
  "TRUE" = Params$ColorHighlight,
  "FALSE" = Params$ColorGray
)

ggplot(
  copy(InitialZoneSimResults)[, `:=`(
    taxi_company = factor(taxi_company, levels = c("Uber", "Lyft")),
    time_hour = reorder(time_hour, time_hour, FUN = sum),
    week_day = factor(Weekdays[week_day], levels = Weekdays)
  )][,
    better_than_model := median(daily_hourly_wage_mean) > FinalPfaMeans[2L],
    by = c("time_hour", "taxi_company", "week_day")
  ],
  aes(
    time_hour,
    daily_hourly_wage_mean,
    fill = better_than_model
  )
) +
  geom_boxplot() +
  geom_hline(yintercept = FinalPfaMeans[1L], linetype = 2) +
  geom_hline(yintercept = FinalPfaMeans[2L], linetype = 2) +
  scale_fill_manual(
    values = BetterThanModelFillColors
  ) +
  scale_y_continuous(breaks = BreaksToPlot, label = dollar_format(1)) +
  facet_grid(vars(taxi_company), vars(week_day)) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**Start Time**</span> ",
      "**Can Increase Earning Reguardless Day of Week**"
    ),
    x = "Hour Of The Day",
    y = "Daily Hourly Wage"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(size = 12, family = Params$BaseFontFamily),
    plot.subtitle = element_markdown(size = 10, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 6, family = Params$BaseFontFamily),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(
      size = 10,
      face = "bold",
      family = Params$BaseFontFamily
    )
  )
```

## Finding hidden patterns

Based on the prior exploration it's clear that the most important rule found so far is to take Uber as the company, but to improve earning we need to find better interactions between variables and for that reason we are going to start with **tree model to see the model can find the most important interactions** to explain changes in the target variable.

To tune the model we know that the must important hyperparam is the `tree_depth` to limit how many splits the tree can make, which controls vertical complexity of the model and as result we want to use the **minimum value** posible.

As we have a lot of data to explore due the number of zones of the experiment we know that the `min_n` to set the minimum number of data points required in a node to allow a further split won't be a important hyperparam as must of the split will have many data points to avoid overfit.

```{r}
DecisionTreeSpec <- decision_tree(
  mode = "classification",
  tree_depth = tune("tree_depth"),
  min_n = 10
)
```

To help the tree finding the important pattern we are going to apply the following transformations to the orininal data:

* As `start_zone` present high cardinality of zone we substitute with `hourly_wage_mean_by_zone` after isolating the effect of taxi company.
* To allow the model to see hour of the day and the time of  the week as the cycles they represent in the real live where the same time is repited very 24 hours and the same time in the week is repeted every 168 hours we are creating the variabes `day_cycle_sin_1` and `day_cycle_cos_1` instead of just showing `time_hour` and `week_cycle_sin_1` and `week_cycle_cos_1` instead of using the original `week_day`.
* We are showing the `taxi_company` company as boolean variable to keep all variable numeric.

```{r}
#| label: Defining Tree Workdflow
DecisionTreeRecipe <- recipe(
  is_top_trip ~ taxi_company +
    start_zone +
    week_day +
    time_hour +
    simulation_id +
    daily_hourly_wage_mean,
  data = InitialZoneSimResults
) |>
  # Creating variables to use later
  step_mutate(
    day_cycle = time_hour,
    week_cycle = (week_day - 1) * 24 + time_hour,
    hourly_wage_mean_by_zone = paste0(start_zone, "-", taxi_company)
  ) |>
  # Changing zones with their mean
  step_lencode(
    hourly_wage_mean_by_zone,
    outcome = vars(daily_hourly_wage_mean),
    smooth = FALSE
  ) |>
  # Transforming company to binary variable
  step_dummy(taxi_company) |>
  # Creating Cycles
  step_harmonic(
    day_cycle,
    cycle_size = 24,
    frequency = 1,
    keep_original_cols = FALSE
  ) |>
  step_harmonic(
    week_cycle,
    cycle_size = 168,
    frequency = 1,
    keep_original_cols = FALSE
  ) |>
  # Saving original vars
  update_role(
    simulation_id,
    daily_hourly_wage_mean,
    start_zone,
    time_hour,
    week_day,
    new_role = "additional info"
  ) |>
  # Removing original variables
  step_upsample(is_top_trip, over_ratio = tune("over_ratio"))
```

Now after all this configuration we are ready to consolidate all that logic into a compact workflow that will help us to fit and evaluate our model using all tools from tidymodels.

```{r}
DecisionTreeWf <-
  workflow() |>
  add_recipe(DecisionTreeRecipe) |>
  add_model(DecisionTreeSpec)
```

As this as simple model and we just need to tune 2 params we can use the grid method the find the best trip possible with 10 fold cross validation to estimate the model error without taking testing data, as we just need to understand the results of the experiment rather than using this model to make predictions.

```{r}
#| eval: false
#| label: Tuning tree model
DecisionTreeWfRegularGrid <- grid_regular(
  tree_depth(range = c(2L, 16L)),
  over_ratio(range = c(0.1, 1)),
  levels = c(tree_depth = 5, over_ratio = 10)
)

set.seed(5878)
InitialZoneSimResultsResamples <- vfold_cv(
  InitialZoneSimResults,
  v = 10,
  strata = is_top_trip
)

DecisionTreeWfTuned <- tune_grid(
  object = DecisionTreeWf,
  resamples = InitialZoneSimResultsResamples,
  grid = DecisionTreeWfRegularGrid,
  metrics = metric_set(recall, precision, brier_class),
  control = control_grid(
    verbose = TRUE,
    parallel_over = NULL,
    save_pred = FALSE
  )
)

pin_write(
  BoardLocal,
  DecisionTreeWfTuned,
  name = "DecisionTreeWfTuned",
  type = "qs2"
)
```

After making all this expliration we not only need to take the configuration that better predict the results, but instead the one that loss less prediction power without moving too high in the `tree_depth`. For that reason rathen than using the default function `select_best` we are going to use a diferent version that is more flexible in the **brier_class** score to select the configuration with less model complexity as we want a model **easy to explain**.

```{r}
DecisionTreeWfTuned <- pin_read(
  BoardLocal,
  name = "DecisionTreeWfTuned"
)

DecisionTreeWfBestConfig <- select_by_pct_loss(
  DecisionTreeWfTuned,
  metric = "brier_class",
  limit = 1,
  tree_depth
)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Results of tunning
autoplot(DecisionTreeWfTuned) +
  geom_vline(xintercept = DecisionTreeWfBestConfig$over_ratio, linetype = 2) +
  scale_x_continuous(breaks = breaks_width(0.1)) +
  labs(
    title = paste0(
      "**With a tree depth of**",
      highlight_text(DecisionTreeWfBestConfig$tree_depth),
      "**and**",
      DecisionTreeWfBestConfig$over_ratio |>
        scales::percent() |>
        highlight_text(),
      "**of oversampling**<br>**We can get the best model**"
    )
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    legend.position = "top",
    plot.title = element_markdown(family = Params$BaseFontFamily),
    plot.subtitle = element_text(
      family = Params$BaseFontFamily,
      face = "italic"
    ),
    legend.title = element_markdown(family = Params$BaseFontFamily),
    panel.grid.minor = element_blank()
  )
```

Now we are ready to train train the model and see the probabilities predicted by the model are related with the variable that we want to maximize and the results we really good as we can see that the median of `daily_hourly_wage_mean` increases with the probability of the trip to be of high value.

```{r}
#| eval: false
DecisionTreeWfFitted <-
  finalize_workflow(DecisionTreeWf, DecisionTreeWfBestConfig) |>
  fit(data = InitialZoneSimResults)

pin_write(
  BoardLocal,
  DecisionTreeWfFitted,
  name = "DecisionTreeWfFitted",
  type = "qs2"
)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Tree prop vs target variable
DecisionTreeWfFitted <- pin_read(
  BoardLocal,
  "DecisionTreeWfFitted"
)

DecisionTreeAugmented <-
  augment(DecisionTreeWfFitted, new_data = InitialZoneSimResults)

ggplot(
  DecisionTreeAugmented,
  aes(
    round(.pred_yes, 2) |> forcats::as_factor(),
    daily_hourly_wage_mean
  )
) +
  geom_jitter(aes(color = is_top_trip), alpha = 0.1) +
  geom_boxplot() +
  geom_hline(yintercept = TopDailyHourlyWage, linetype = 2) +
  scale_y_continuous(
    breaks = seq(0, 150, 10) |> setdiff(70) |> c(unname(TopDailyHourlyWage)),
    label = label_currency(accuracy = 1)
  ) +
  scale_color_manual(
    values = c("yes" = Params$ColorHighlight, "no" = Params$ColorGray)
  ) +
  labs(
    title = paste0(
      "**After**",
      highlight_text("50%"),
      "**we can confirm a change in distribution**"
    ),
    y = "Daily Hourly Wage",
    x = "Probabilty to be High Trip"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(family = Params$BaseFontFamily),
    axis.title = element_text(size = 12, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(
      family = Params$BaseFontFamily,
      face = "italic"
    ),
    legend.position = "none"
  )
```

I feel really good of the resolts provided in the confution matrix, as it was able to predict 41% of days with high value and only predict as high value trips wrongly for 5% of all trips possible, which seems to be a solic pattern. 

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Conf matrix
plot_custom_conf_matrix <- function(dt) {
  procced_data =
    dt |>
    group_by(Truth = is_top_trip, Prediction = .pred_class) |>
    summarize(Freq = n(), mean_pred_yes = mean(.pred_yes)) |>
    mutate(truth_prop = Freq / sum(Freq)) |>
    ungroup()

  true_positive_prob =
    procced_data |>
    filter(Truth == "yes", Prediction == "yes") |>
    pull(truth_prop) |>
    percent()

  created_plot =
    procced_data |>
    ggplot(aes(Truth, Prediction)) +
    geom_tile(aes(fill = truth_prop)) +
    geom_label(
      aes(
        label = paste0(
          percent(truth_prop),
          "\n",
          comma(Freq),
          "\n Yes prob: ",
          percent(mean_pred_yes)
        )
      ),
      fill = "white"
    ) +
    scale_fill_continuous(palette = "Greys") +
    labs(
      title = paste0(
        true_positive_prob |> paste("of high value trips") |> highlight_text(),
        "**were predicted**"
      )
    ) +
    theme_minimal(base_family = Params$BaseFontFamily) +
    theme(
      plot.title.position = "plot",
      plot.title = element_markdown(family = Params$BaseFontFamily),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(size = 12),
      panel.grid = element_blank(),
      legend.position = 'none'
    )

  return(created_plot)
}

plot_custom_conf_matrix(DecisionTreeAugmented)
```


## Confirming pattern effect

To facilite the interpretation of the effect of using the tree model to change the behavior of the taxi driver, we will range all variables from 0 to 1 in under to have the model coefficients in dollars per hour.

```{r}
TreeResultsToExplore <-
  DecisionTreeRecipe |>
  step_range(all_numeric_predictors()) |>
  finalize_recipe(DecisionTreeWfBestConfig) |>
  prep(training = InitialZoneSimResults) |>
  bake(new_data = InitialZoneSimResults) |>
  bind_cols(
    transmute(
      DecisionTreeAugmented,
      tree_model_high_value = (.pred_yes > 0.5) * 1
    )
  )
```

As we want to define the effect of using the tree model over the hourly mean wage of the drivers it's really important to confirm that the variables used to train the model don't present colliniality and based on the table it's clear that `taxi_company_Uber` and `hourly_wage_mean_by_zone` are apporting the same information. As result we know that `hourly_wage_mean_by_zone` it's just dividing drivers's company an **the zone is not providing any information.**

```{r}
linear_reg() |>
  set_engine("lm") |>
  fit(
    data = TreeResultsToExplore,
    formula = daily_hourly_wage_mean ~
      tree_model_high_value +
      hourly_wage_mean_by_zone +
      taxi_company_Uber +
      day_cycle_sin_1 +
      day_cycle_cos_1 +
      week_cycle_sin_1 +
      week_cycle_cos_1
  ) |>
  extract_fit_engine() |>
  car::vif() |>
  round(digits = 2) |>
  (\(x) tibble(variable = names(x), vif = x))() |>
  arrange(desc(vif)) |>
  knitr::kable()
```

With 65,856 simulated days all coefficients are statistically significant,
so the focus should be on the magnitude of each effect.

The regression reveals a clear hierarchy of factors:

- A driver that use Uber
  can earn up to **$8.16 more per hour** compared to a driver that use Lyft.
- Starting at a favorable time of day can add up to **$5.08 per hour** (captured by `day_cycle_cos_1`), while an unfavorable hour can subtract up to **$2.94**.
- Day-of-week effects are small (under **$1.58** in either direction),
  confirming that *what time of day* the driver starts matters far more
  than *which day of the week* they work.
- A driver who follows the tree model's accept/reject strategy earns an
  additional **$5.04 per hour** on top of all the above. This reflects
  the specific company-and-timing combinations the tree identified as
  high-value that **no single variable captures independently**.

In practice, a driver making all the right decisions — starting with Uber 
at a favorable time and following the tree model's 
strategy — can move from the baseline of **$54/hour** to approximately 
**$70/hour**, representing a **`r round((8.16+5.08+5.04)/53.83,2)*100`% improvement** in hourly earnings.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: lm coef table
create_lm_gt_table = function(lm_model) {
  model_coef =
    tidy(lm_model, conf.int = TRUE) |>
    arrange(desc(abs(estimate))) |>
    select(term, conf.low, estimate, conf.high)

  gt_final =
    gt(model_coef) |>
    tab_header(
      title = "Impact Analysis: Decision Tree Policy vs. Baseline Components",
      subtitle = "Hierarchy of factors determining hourly income"
    ) |>
    cols_label(
      term = "Variable",
      conf.low = "Lower",
      estimate = "Estimate (Coeff.)",
      conf.high = "Upper"
    ) |>
    fmt_currency(columns = c(conf.low, estimate, conf.high), decimals = 2) |>
    tab_style(
      style = cell_text(style = "italic"),
      locations = cells_body(columns = term)
    ) |>
    tab_style(
      style = list(
        cell_fill(color = Params$ColorHighlightLow),
        cell_text(weight = "bold")
      ),
      locations = cells_body(rows = term == "tree_model_high_value")
    ) |>
    tab_source_note(
      source_note = "Note: Based on 65,856 simulated days. OLS regression with robust standard errors."
    )

  return(gt_final)
}

LinearModelTreeEffect <- linear_reg() |>
  set_engine("lm") |>
  fit(
    data = TreeResultsToExplore,
    formula = daily_hourly_wage_mean ~
      tree_model_high_value +
      taxi_company_Uber +
      day_cycle_sin_1 +
      day_cycle_cos_1 +
      week_cycle_sin_1 +
      week_cycle_cos_1
  )

create_lm_gt_table(LinearModelTreeEffect)
```

## Extracting bussines rules from tree model

To understand how the model can get so incredible results let's just plot the model to see how it is using the features.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Tree Plot
#| fig-height: 7
TreeModelFittedEngine = extract_fit_engine(DecisionTreeWfFitted)

rpart.plot::prp(
  # 1. Model Object
  x = TreeModelFittedEngine,

  # 2. Labels & Formatting
  main = "Strategy to find best trips",
  prefix = "Best trip\n",
  extra = 106,
  digits = 2,

  # 3. Tree Structure & Geometry
  type = 3,
  branch = 0.3,
  branch.lwd = 2.5,
  roundint = FALSE,

  # 4. Node & Leaf Aesthetics
  box.palette = "GyPu",
  round = 2,
  leaf.round = 9,
  under = TRUE,
  under.cex = 1.0,

  # 5. Variable Name Handling
  varlen = 0,
  faclen = 0,
  clip.facs = FALSE,

  # 6. Typography & Fonts
  family = Params$BaseFontFamily,
  split.family = Params$BaseFontFamily,
  split.cex = 0.9
)
```

As the must important variable to for the model was the `hourly_wage_mean_by_zone` let's explore which zones were excluded from the zones for the model. As first step, we can extract the split used for that variable.

```{r}
HourlyWageMeanByZoneSplit <-
  TreeModelFittedEngine$splits["hourly_wage_mean_by_zone", "index"]

HourlyWageMeanByZoneSplit
```

As the variable used by model are not easy to be interpreted by humans let's integrate in a sigle dataframe the features used by the model, the features before using the recipe transformations and the preductions of the model to create visuals that translate what was learn by the model.

```{r}
TreeResultsToExplain <-
  DecisionTreeRecipe |>
  finalize_recipe(DecisionTreeWfBestConfig) |>
  prep(training = InitialZoneSimResults) |>
  bake(new_data = InitialZoneSimResults) |>
  (\(df) {
    select(df, -all_of(intersect(names(df), names(DecisionTreeAugmented))))
  })() |>
  bind_cols(
    DecisionTreeAugmented
  )
```


1. The `hourly_wage_mean_by_zone` (as explained before) only worked to seprate trips between **Uber** and **Lyft** to only select **Uber** at any zone.

```{r}
TreeResultsToExplain |>
  mutate(is_top_zone = hourly_wage_mean_by_zone >= HourlyWageMeanByZoneSplit) |>
  distinct(start_zone, is_top_zone) |>
  count(start_zone) |>
  filter(n == 1L) |>
  nrow()
```

2. It's better to start working at night in order to get better results.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Hour Tree Explanation
TreeResultsToExplain |>
  filter(taxi_company_Uber == 1L) |>
  distinct(
    .pred_class,
    time_hour,
    day_cycle_sin_1,
    day_cycle_cos_1,
    week_day = factor(
      Weekdays[week_day],
      levels = c(Weekdays[-1L], Weekdays[1L])
    ),
    week_cycle_sin_1,
    week_cycle_cos_1
  ) |>
  ggplot(aes(
    day_cycle_sin_1,
    day_cycle_cos_1,
    color = .pred_class
  )) +
  geom_text(aes(label = time_hour), fontface = "bold", size = 3.5) +
  scale_color_manual(
    values = c("yes" = Params$ColorHighlight, "no" = Params$ColorGray)
  ) +
  facet_wrap(vars(week_day), ncol = 4) +
  coord_fixed() +
  labs(
    title = paste0(
      "**Must of the days taxi driver can start at**",
      highlight_text("21 hours")
    ),
    subtitle = "_On **Mondays** taxi drivers could just rest_"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily, base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_rect(color = "black"),
    strip.text = element_text(face = "bold", size = 12),
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title = element_markdown(),
    plot.subtitle = element_markdown()
  )
```

