---
title: "From Predictions to Policies: Integrating ML into Stochastic Optimization"
subtitle: ""
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once have a policy to decide if a trip should be taken or rejected then it's important to validate if any of the next factors could cause an increase of decrease on taxi results after blocking the **company** for taxi driver:

- Start zone
- Days of the week
- Hour of the day

## Defining combinations to simulate

As we have 196 zones to evaluete, 3 days to evaluete for each day of the week (7), 2 taxi companies and we divide the day in 8 moments to measure (one each 3 hours), we end with an experiment of **65,856 days** to simulate.

```{r}
#| label: Defining combinations
# Loading functions ----

## To manage relative paths
library(here)
library(pins)
library(qs2)

## Manipulate data
library(data.table)
library(DBI)
library(duckdb)
library(lubridate)
library(timeDate)
library(ggtext)

## To run predictions
library(tidymodels)
library(embed)
library(themis)

## Custom functions, board & params
devtools::load_all()
BoardLocal <- board_folder(here("../NycTaxiPins/Board"))
Params <- yaml::read_yaml(here("params.yml"))


# Defining combinations ----

## Getting shapes to use
ZonesShapes <-
  pin_read(BoardLocal, "ZonesShapes") |>
  as.data.frame() |>
  subset(
    borough %in% names(Params$BoroughColors),
    select = c("LocationID", "borough", "zone")
  )

setDT(ZonesShapes)
ZonesShapes[, LocationID := as.character(LocationID)]

UniqueZoneIds <- unique(ZonesShapes$LocationID) |> sort()


## Defining days to sample
YearDays <- data.table(
  days = seq.Date(make_date(2023), make_date(2023, 12, 31), by = "day")
)
YearDays[, wday := lubridate::wday(days)]
set.seed(4879)
SampledDays <- YearDays[, .SD[sample.int(.N, 3L)], by = "wday"]


## Pre-build the big design matrix once
DataToSimulate <- CJ(
  DOLocationID = UniqueZoneIds,
  hvfhs_license_num = c("HV0003", "HV0005"),
  calendar_day = as.character(SampledDays$days),
  time_hour = seq(0, 21, length.out = 8)
)
DataToSimulate[, `:=`(
  trip_id = as.numeric(1:.N),
  request_datetime = paste(calendar_day, time_hour) |> ymd_h(),
  trip_time = 0
)]

head(DataToSimulate, 8)
```

## Running simulation

As I knew that running the hole process would take me several days I decided to store the results of simulations on disk for each zone. In this way I was able to keep the experiment as I was expecting and stop the process between days to allow the computer to rest between days and redice the risk of loossing all the progress I had problems with memory while using the computer for dealy activities.

To avoid the overhead of using a IDE the next code was source from terminal after connecting with `ssh NycTaxi` to consume the fewer results as possible.

::: {.callout-note}
In the below code you see an assigning the starting zone to `DOLocationID` as the simulation uses the last zone of the configuration trip (the one we are creting) to starting picking trips from that zone.
:::

```{r}
#| eval: false
#| label: Simulation code
## Loading env

# Getting prior policy
AcceptRejectPolicyFitted <-
  pin_read(BoardLocal, "AcceptRejectPolicyFitted")

# Competing renaming columns
DataToSimulate[, `:=`(
  calendar_day = NULL,
  time_hour = NULL,
  PULocationID = "1",
  wav_match_flag = "N",
  driver_pay = 0,
  tips = 0,
  dispatching_base_num = "",
  originating_base_num = "",
  on_scene_datetime = NA_POSIXct_,
  pickup_datetime = NA_POSIXct_,
  dropoff_datetime = NA_POSIXct_,
  trip_miles = 0,
  base_passenger_fare = 0,
  tolls = 0,
  bcf = 0,
  sales_tax = 0,
  congestion_surcharge = 0,
  airport_fee = 0,
  shared_request_flag = "",
  shared_match_flag = "",
  access_a_ride_flag = "",
  wav_request_flag = "",
  month = 1,
  year = 2023,
  performance_per_hour = 0
)]

# Check which zones are already done
AlreadyDone <- pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  gsub(pattern = "InitialZoneSim_ZoneId_", replacement = "")

zones_to_run <- setdiff(UniqueZoneIds, AlreadyDone)

cat(
  "→",
  length(zones_to_run),
  "zones left to simulate (out of",
  length(UniqueZoneIds),
  "total).\n"
)


# === MAIN RESUMABLE LOOP ===
for (zone_i in zones_to_run) {
  cat("   Simulating zone", zone_i, "... ")

  # Filter only this zone
  ZoneData <- DataToSimulate[list(zone_i), on = "DOLocationID"]

  # Fresh connection for each zone
  con <- dbConnect(duckdb(), dbdir = here("../NycTaxiBigFiles/my-db.duckdb"))

  ZoneSim <- simulate_trips(
    con,
    ZoneData,
    seeds = 4878,
    fitted_wf = AcceptRejectPolicyFitted
  )

  dbDisconnect(con, shutdown = TRUE)

  # Save immediately with clear name
  pin_write(
    BoardLocal,
    ZoneSim,
    paste0("InitialZoneSim_ZoneId_", zone_i),
    type = "qs2"
  )

  cat("DONE (", nrow(ZoneSim), "rows)\n")
}

cat("✅ All zones simulated!\n")
```

Here is the list of zone simulated and we can see that it took around **102 minutes** to complete each zone which mean that I invest around **14 days** of computation to see the next results.

```bash
angelfeliz@debian:~/r-projects/NycTaxi$ ls -lahd ../NycTaxiPins/Board/InitialZoneSim_ZoneId_*
drwxr-xr-x 3 root root 4.0K Apr  7 10:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_10
drwxr-xr-x 3 root root 4.0K Apr  7 12:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_100
drwxr-xr-x 3 root root 4.0K Apr  7 13:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_101
drwxr-xr-x 3 root root 4.0K Apr  7 15:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_102
drwxr-xr-x 3 root root 4.0K Apr  7 17:06 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_103
drwxr-xr-x 3 root root 4.0K Apr  7 18:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_106
drwxr-xr-x 3 root root 4.0K Apr  7 20:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_107
drwxr-xr-x 3 root root 4.0K Apr  7 22:31 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_108
drwxr-xr-x 3 root root 4.0K Apr  8 00:11 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_11
drwxr-xr-x 3 root root 4.0K Apr  8 01:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_111
drwxr-xr-x 3 root root 4.0K Apr  8 03:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_112
drwxr-xr-x 3 root root 4.0K Apr  8 05:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_113
drwxr-xr-x 3 root root 4.0K Apr  8 06:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_114
drwxr-xr-x 3 root root 4.0K Apr  8 08:35 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_116
drwxr-xr-x 3 root root 4.0K Apr  8 10:16 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_117
drwxr-xr-x 3 root root 4.0K Apr  8 11:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_12
drwxr-xr-x 3 root root 4.0K Apr  8 13:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_120
drwxr-xr-x 3 root root 4.0K Apr  8 15:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_121
drwxr-xr-x 3 root root 4.0K Apr  8 16:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_122
drwxr-xr-x 3 root root 4.0K Apr  8 18:40 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_123
drwxr-xr-x 3 root root 4.0K Apr  8 20:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_124
drwxr-xr-x 3 root root 4.0K Apr  8 22:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_125
drwxr-xr-x 3 root root 4.0K Apr  9 01:50 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_127
drwxr-xr-x 3 root root 4.0K Apr  9 03:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_128
drwxr-xr-x 3 root root 4.0K Apr  9 05:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_129
drwxr-xr-x 3 root root 4.0K Apr  9 06:51 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_13
drwxr-xr-x 3 root root 4.0K Apr  9 08:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_130
drwxr-xr-x 3 root root 4.0K Apr  9 11:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_131
drwxr-xr-x 3 root root 4.0K Apr  9 13:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_132
drwxr-xr-x 3 root root 4.0K Apr  9 15:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_133
drwxr-xr-x 3 root root 4.0K Apr  9 16:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_134
drwxr-xr-x 3 root root 4.0K Apr  9 18:23 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_135
drwxr-xr-x 3 root root 4.0K Apr  9 20:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_137
drwxr-xr-x 3 root root 4.0K Apr  9 21:51 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_138
drwxr-xr-x 3 root root 4.0K Apr  9 23:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_139
drwxr-xr-x 3 root root 4.0K Apr 10 01:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_14
drwxr-xr-x 3 root root 4.0K Apr 10 03:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_140
drwxr-xr-x 3 root root 4.0K Apr 10 05:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_141
drwxr-xr-x 3 root root 4.0K Apr 10 06:41 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_142
drwxr-xr-x 3 root root 4.0K Apr 10 10:25 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_143
drwxr-xr-x 3 root root 4.0K Apr 10 12:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_144
drwxr-xr-x 3 root root 4.0K Apr 10 13:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_145
drwxr-xr-x 3 root root 4.0K Apr 10 15:33 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_146
drwxr-xr-x 3 root root 4.0K Apr 10 17:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_148
drwxr-xr-x 3 root root 4.0K Apr 10 18:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_149
drwxr-xr-x 3 root root 4.0K Apr 10 20:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_15
drwxr-xr-x 3 root root 4.0K Apr 10 22:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_150
drwxr-xr-x 3 root root 4.0K Apr 11 00:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_151
drwxr-xr-x 3 root root 4.0K Apr 11 01:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_152
drwxr-xr-x 3 root root 4.0K Apr 11 03:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_153
drwxr-xr-x 3 root root 4.0K Apr 11 05:02 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_154
drwxr-xr-x 3 root root 4.0K Apr 11 06:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_155
drwxr-xr-x 3 root root 4.0K Apr 11 08:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_157
drwxr-xr-x 3 root root 4.0K Apr 11 10:07 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_158
drwxr-xr-x 3 root root 4.0K Apr 11 11:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_16
drwxr-xr-x 3 root root 4.0K Apr 11 13:47 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_160
drwxr-xr-x 3 root root 4.0K Apr 11 15:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_161
drwxr-xr-x 3 root root 4.0K Apr 12 00:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_162
drwxr-xr-x 3 root root 4.0K Apr 12 02:32 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_163
drwxr-xr-x 3 root root 4.0K Apr 12 04:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_164
drwxr-xr-x 3 root root 4.0K Apr 12 05:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_165
drwxr-xr-x 3 root root 4.0K Apr 12 07:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_166
drwxr-xr-x 3 root root 4.0K Apr 12 09:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_17
drwxr-xr-x 3 root root 4.0K Apr 12 10:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_170
drwxr-xr-x 3 root root 4.0K Apr 12 12:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_171
drwxr-xr-x 3 root root 4.0K Apr 12 14:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_173
drwxr-xr-x 3 root root 4.0K Apr 12 16:13 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_175
drwxr-xr-x 3 root root 4.0K Apr 12 18:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_177
drwxr-xr-x 3 root root 4.0K Apr 12 20:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_178
drwxr-xr-x 3 root root 4.0K Apr 12 22:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_179
drwxr-xr-x 3 root root 4.0K Apr 12 23:49 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_180
drwxr-xr-x 3 root root 4.0K Apr 13 01:33 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_181
drwxr-xr-x 3 root root 4.0K Apr 13 03:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_186
drwxr-xr-x 3 root root 4.0K Apr 13 04:56 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_188
drwxr-xr-x 3 root root 4.0K Apr 13 06:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_189
drwxr-xr-x 3 root root 4.0K Apr 13 08:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_19
drwxr-xr-x 3 root root 4.0K Apr 13 09:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_190
drwxr-xr-x 3 root root 4.0K Apr 13 11:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_191
drwxr-xr-x 3 root root 4.0K Apr 13 13:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_192
drwxr-xr-x 3 root root 4.0K Apr 13 14:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_193
drwxr-xr-x 3 root root 4.0K Apr 13 18:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_194
drwxr-xr-x 3 root root 4.0K Apr 13 19:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_195
drwxr-xr-x 3 root root 4.0K Apr 13 21:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_196
drwxr-xr-x 3 root root 4.0K Apr 13 23:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_197
drwxr-xr-x 3 root root 4.0K Apr 14 01:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_198
drwxr-xr-x 3 root root 4.0K Apr 14 02:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_2
drwxr-xr-x 3 root root 4.0K Apr 14 04:35 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_201
drwxr-xr-x 3 root root 4.0K Apr 14 06:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_202
drwxr-xr-x 3 root root 4.0K Apr 14 07:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_203
drwxr-xr-x 3 root root 4.0K Apr 14 09:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_205
drwxr-xr-x 3 root root 4.0K Apr 14 11:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_207
drwxr-xr-x 3 root root 4.0K Apr 14 13:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_209
drwxr-xr-x 3 root root 4.0K Apr 14 14:44 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_21
drwxr-xr-x 3 root root 4.0K Apr 14 16:27 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_210
drwxr-xr-x 3 root root 4.0K Apr 14 18:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_211
drwxr-xr-x 3 root root 4.0K Apr 14 22:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_215
drwxr-xr-x 3 root root 4.0K Apr 15 00:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_216
drwxr-xr-x 3 root root 4.0K Apr 15 02:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_217
drwxr-xr-x 3 root root 4.0K Apr 15 04:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_218
drwxr-xr-x 3 root root 4.0K Apr 15 05:56 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_219
drwxr-xr-x 3 root root 4.0K Apr 15 07:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_22
drwxr-xr-x 3 root root 4.0K Apr 15 09:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_222
drwxr-xr-x 3 root root 4.0K Apr 15 11:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_223
drwxr-xr-x 3 root root 4.0K Apr 15 13:02 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_224
drwxr-xr-x 3 root root 4.0K Apr 15 14:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_225
drwxr-xr-x 3 root root 4.0K Apr 15 16:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_226
drwxr-xr-x 3 root root 4.0K Apr 15 18:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_227
drwxr-xr-x 3 root root 4.0K Apr 15 19:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_228
drwxr-xr-x 3 root root 4.0K Apr 15 21:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_229
drwxr-xr-x 3 root root 4.0K Apr 15 23:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_230
drwxr-xr-x 3 root root 4.0K Apr 16 00:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_231
drwxr-xr-x 3 root root 4.0K Apr 16 02:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_232
drwxr-xr-x 3 root root 4.0K Apr 16 04:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_233
drwxr-xr-x 3 root root 4.0K Apr 16 05:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_234
drwxr-xr-x 3 root root 4.0K Apr 16 07:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_236
drwxr-xr-x 3 root root 4.0K Apr 16 09:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_237
drwxr-xr-x 3 root root 4.0K Apr 16 11:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_238
drwxr-xr-x 3 root root 4.0K Apr 16 12:45 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_239
drwxr-xr-x 3 root root 4.0K Apr 16 14:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_24
drwxr-xr-x 3 root root 4.0K Apr 16 16:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_243
drwxr-xr-x 3 root root 4.0K Apr 16 17:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_244
drwxr-xr-x 3 root root 4.0K Apr 16 19:25 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_246
drwxr-xr-x 3 root root 4.0K Apr 16 21:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_249
drwxr-xr-x 3 root root 4.0K Apr 16 22:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_25
drwxr-xr-x 3 root root 4.0K Apr 17 00:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_252
drwxr-xr-x 3 root root 4.0K Apr 17 02:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_253
drwxr-xr-x 3 root root 4.0K Apr 17 04:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_255
drwxr-xr-x 3 root root 4.0K Apr 17 05:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_256
drwxr-xr-x 3 root root 4.0K Apr 17 07:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_257
drwxr-xr-x 3 root root 4.0K Apr 17 09:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_258
drwxr-xr-x 3 root root 4.0K Apr 17 11:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_26
drwxr-xr-x 3 root root 4.0K Apr 17 13:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_260
drwxr-xr-x 3 root root 4.0K Apr 17 14:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_261
drwxr-xr-x 3 root root 4.0K Apr 17 16:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_262
drwxr-xr-x 3 root root 4.0K Apr 17 18:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_263
drwxr-xr-x 3 root root 4.0K Apr 17 19:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_27
drwxr-xr-x 3 root root 4.0K Apr 17 21:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_28
drwxr-xr-x 3 root root 4.0K Apr 17 23:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_29
drwxr-xr-x 3 root root 4.0K Apr 18 00:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_30
drwxr-xr-x 3 root root 4.0K Apr 18 02:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_33
drwxr-xr-x 3 root root 4.0K Apr 18 04:16 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_34
drwxr-xr-x 3 root root 4.0K Apr 18 05:55 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_35
drwxr-xr-x 3 root root 4.0K Apr 18 07:40 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_36
drwxr-xr-x 3 root root 4.0K Apr 18 09:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_37
drwxr-xr-x 3 root root 4.0K Apr 18 11:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_38
drwxr-xr-x 3 root root 4.0K Apr 18 13:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_39
drwxr-xr-x 3 root root 4.0K Apr 18 15:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_4
drwxr-xr-x 3 root root 4.0K Apr 18 16:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_40
drwxr-xr-x 3 root root 4.0K Apr 18 19:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_41
drwxr-xr-x 3 root root 4.0K Apr 18 20:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_42
drwxr-xr-x 3 root root 4.0K Apr 18 22:41 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_43
drwxr-xr-x 3 root root 4.0K Apr 19 00:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_45
drwxr-xr-x 3 root root 4.0K Apr 19 02:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_48
drwxr-xr-x 3 root root 4.0K Apr 19 03:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_49
drwxr-xr-x 3 root root 4.0K Apr 19 05:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_50
drwxr-xr-x 3 root root 4.0K Apr 19 07:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_52
drwxr-xr-x 3 root root 4.0K Apr 19 09:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_53
drwxr-xr-x 3 root root 4.0K Apr 19 10:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_54
drwxr-xr-x 3 root root 4.0K Apr 19 12:29 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_55
drwxr-xr-x 3 root root 4.0K Apr 19 14:11 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_56
drwxr-xr-x 3 root root 4.0K Apr 19 16:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_61
drwxr-xr-x 3 root root 4.0K Apr 19 17:50 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_62
drwxr-xr-x 3 root root 4.0K Apr 19 19:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_63
drwxr-xr-x 3 root root 4.0K Apr 19 21:27 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_64
drwxr-xr-x 3 root root 4.0K Apr 19 23:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_65
drwxr-xr-x 3 root root 4.0K Apr 20 01:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_66
drwxr-xr-x 3 root root 4.0K Apr 20 02:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_67
drwxr-xr-x 3 root root 4.0K Apr 20 04:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_68
drwxr-xr-x 3 root root 4.0K Apr 20 06:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_7
drwxr-xr-x 3 root root 4.0K Apr 20 07:46 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_70
drwxr-xr-x 3 root root 4.0K Apr 20 09:31 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_71
drwxr-xr-x 3 root root 4.0K Apr 20 11:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_72
drwxr-xr-x 3 root root 4.0K Apr 20 12:49 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_73
drwxr-xr-x 3 root root 4.0K Apr 20 14:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_74
drwxr-xr-x 3 root root 4.0K Apr 20 16:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_75
drwxr-xr-x 3 root root 4.0K Apr 20 18:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_76
drwxr-xr-x 3 root root 4.0K Apr 20 22:13 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_77
drwxr-xr-x 3 root root 4.0K Apr 21 00:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_79
drwxr-xr-x 3 root root 4.0K Apr 21 02:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_8
drwxr-xr-x 3 root root 4.0K Apr 21 09:47 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_80
drwxr-xr-x 3 root root 4.0K Apr 21 11:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_82
drwxr-xr-x 3 root root 4.0K Apr 21 13:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_83
drwxr-xr-x 3 root root 4.0K Apr 21 17:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_85
drwxr-xr-x 3 root root 4.0K Apr 21 18:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_86
drwxr-xr-x 3 root root 4.0K Apr 21 20:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_87
drwxr-xr-x 3 root root 4.0K Apr 21 23:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_88
drwxr-xr-x 3 root root 4.0K Apr 22 00:45 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_89
drwxr-xr-x 3 root root 4.0K Apr 22 02:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_9
drwxr-xr-x 3 root root 4.0K Apr 22 04:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_90
drwxr-xr-x 3 root root 4.0K Apr 22 05:44 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_91
drwxr-xr-x 3 root root 4.0K Apr 22 07:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_92
drwxr-xr-x 3 root root 4.0K Apr 22 09:26 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_93
drwxr-xr-x 3 root root 4.0K Apr 22 11:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_95
drwxr-xr-x 3 root root 4.0K Apr 22 12:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_96
drwxr-xr-x 3 root root 4.0K Apr 22 14:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_97
drwxr-xr-x 3 root root 4.0K Apr 22 16:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_98
```

Now we just need to summary the results to start the exploration.

```{r}
#| eval: false
#| label: Consolidating results
consolidate_all_simulations <- function(zone_id_data, board, sim_start_days) {
  consolidated_data =
    sim_start_trip_summary(
      pin_read(board, zone_id_data),
      sim_start_days
    )[,
      sim_start_days[, list(
        simulation_id = trip_id,
        taxi_company = fifelse(hvfhs_license_num == 'HV0003', "Uber", "Lyft"),
        start_zone = DOLocationID,
        week_day = wday(calendar_day),
        time_hour
      )][.SD, on = "simulation_id"],
      .SDcols = !patterns("_sd")
    ]

  return(consolidated_data)
}

InitialZoneSimResults <-
  pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  lapply(
    FUN = consolidate_all_simulations,
    board = BoardLocal,
    sim_start_days = DataToSimulate
  ) |>
  rbindlist()

# Adding region info
InitialZoneSimResults[
  ZonesShapes,
  on = c("start_zone" = "LocationID"),
  `:=`(borough = borough, start_zone_name = zone)
]

# Save immediately with clear name
pin_write(
  BoardLocal,
  InitialZoneSimResults,
  "InitialZoneSimResults",
  type = "qs2"
)
```


## Exploring results

Before anything, it's important to confirm if after changing the initial conditions of trips had an effect over the target variable. Just imagene that the we see the same value for all combinations we could stop the question just by making the first chart.

But you can see floowing chart preset a lot variation, it seems like the data is following a **log-normal** due it's right tail. This are really promessing results, the goal behind this experiment is to indentify the starting configurations that can beat the current baseline of **`r here::here("../NycTaxiPins/Board") |> pins::board_folder() |> pins::pin_read( "BaseLineHourlyWage") |> scales::dollar()`** and it's clear based on the chart that the **top 10%** presented higher daily hourly have than the base line.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing distribution
InitialZoneSimResults <- pin_read(
  BoardLocal,
  "InitialZoneSimResults"
)

FinalPfaMeans <- pin_read(
  BoardLocal,
  "FinalPfaMeans"
)


TopDailyHourlyWage <- quantile(
  InitialZoneSimResults$daily_hourly_wage_mean,
  0.90
)

InitialZoneSimResults[,
  is_top_trip := fifelse(
    daily_hourly_wage_mean >= TopDailyHourlyWage,
    "yes",
    "no"
  ) |>
    factor(levels = c("no", "yes"))
]


BreaksToPlot <- c(seq(20, 200, 10), unname(FinalPfaMeans)) |> setdiff(60L)

highlight_text <- function(x) {
  paste0(
    " <span style='color:",
    Params$ColorHighlight,
    ";'>**",
    x,
    "**</span> "
  )
}

ggplot(
  InitialZoneSimResults,
  aes(daily_hourly_wage_mean, fill = is_top_trip)
) +
  geom_histogram(color = "black", bins = 30) +
  geom_vline(
    xintercept = FinalPfaMeans[2L],
    linetype = 2
  ) +
  scale_y_continuous(
    label = scales::label_comma(accuracy = 1)
  ) +
  scale_x_continuous(
    label = scales::label_currency(accuracy = 1),
    breaks = setdiff(BreaksToPlot, FinalPfaMeans[1L])
  ) +
  scale_fill_manual(
    values = c("yes" = Params$ColorHighlight, "no" = Params$ColorGray)
  ) +
  labs(
    title = paste0(
      "**Top**",
      highlight_text("10%"),
      "**of combinations make at least**",
      highlight_text("_$70 per hour_")
    ),
    y = "Count",
    x = "Daily Hourly Wage"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily, base_size = 12) +
  theme(
    plot.title = element_markdown(size = 14, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```

As we can see the improment behind the the policy only benefit a **Uber** drivers as the median Hourly Wage for **Lyft** drivers match this the exact value of the baseline, as result our policy will always recomend **Uber** as taxi company.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing taxi company effect
TaxiCompanyFillColors <- c(
  "Uber" = Params$ColorHighlight,
  "Lyft" = Params$ColorGray
)

ggplot(
  InitialZoneSimResults,
  aes(
    factor(taxi_company, levels = names(TaxiCompanyFillColors)),
    daily_hourly_wage_mean,
    fill = taxi_company
  )
) +
  geom_boxplot(width = 0.5) +
  geom_hline(
    yintercept = FinalPfaMeans[2L],
    linetype = 2
  ) +
  geom_hline(
    yintercept = FinalPfaMeans[1L],
    linetype = 2
  ) +
  scale_fill_manual(
    values = TaxiCompanyFillColors
  ) +
  scale_y_continuous(
    breaks = BreaksToPlot,
    label = dollar_format(1)
  ) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**_Uber_**</span> ",
      "**drivers makes more money than** ",
      "<span style='color:",
      "#5b5b5b",
      ";'>**_Lyft_**</span> **drivers**"
    ),
    x = "Taxi Company",
    y = "Daily Hourly Wage"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(size = 14, family = Params$BaseFontFamily),
    plot.subtitle = element_markdown(size = 12, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  )
```

After isoliating the effect of the taxi company in the hourly wage we can see by selected the zones with highest and lowers median that change in the target variable explanied by starting zone is low due that that we are evaluating **`{r} uniqueN(InitialZoneSimResults$start_zone)` zones** that increase in the median of Uber drivers was only **$2** over or baseline ($`{r} round(FinalPfaMeans[2L], 2)`) after implementing the trip selection model and based on the boxplot we can conclude that zones don't provide enough information to explain the variance present in the target variable.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing zone effect
TopBottomZonesMedian <- InitialZoneSimResults[,
  list(daily_hourly_wage_mean = median(daily_hourly_wage_mean)),
  by = c("taxi_company", "start_zone")
][order(daily_hourly_wage_mean)][,
  .SD[c(1:10, .N - 0:9)],
  by = "taxi_company",
  .SDcols = "start_zone"
][, InitialZoneSimResults[.SD, on = c("taxi_company", "start_zone")]][,
  start_zone_name := reorder(
    paste0(start_zone_name, "-", taxi_company),
    daily_hourly_wage_mean,
    median
  )
]

ggplot(
  TopBottomZonesMedian,
  aes(daily_hourly_wage_mean, start_zone_name, fill = borough)
) +
  geom_boxplot() +
  geom_vline(xintercept = FinalPfaMeans[1L], linetype = 2) +
  geom_vline(xintercept = FinalPfaMeans[2L], linetype = 2) +
  scale_fill_manual(values = Params$BoroughColors) +
  scale_x_continuous(
    breaks = BreaksToPlot,
    label = scales::dollar_format(1)
  ) +
  scale_y_discrete(label = \(x) sub("-(.|\\s)+$", "", x)) +
  facet_wrap(vars(taxi_company), ncol = 1, scales = "free_y") +
  labs(
    title = "Some Difference Can Be Explained Start Zone",
    y = "Start Simulation Zone",
    x = "Daily Hourly Wage",
    fill = "Borough"
  ) +
  theme_minimal(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold",
      family = Params$BaseFontFamily
    ),
    axis.text.y = element_text(
      size = 10,
      face = "italic",
      family = Params$BaseFontFamily
    ),
    axis.title = element_text(face = "bold", family = Params$BaseFontFamily),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold", family = Params$BaseFontFamily),
    strip.text = element_text(
      size = 12,
      face = "bold",
      family = Params$BaseFontFamily
    )
  )
```

Based on the next chart we can see a strong effect of start hour in the target variable and some moderate interactions with the day of the week as the hourly wage is higher at 0 hours on Sundays and Saturdays, but those increases are not present on week days.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: Showing hour cycles effect
Weekdays <- c(
  "Sunday",
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday"
)

BetterThanModelFillColors <- c(
  "TRUE" = Params$ColorHighlight,
  "FALSE" = Params$ColorGray
)

ggplot(
  copy(InitialZoneSimResults)[, `:=`(
    taxi_company = factor(taxi_company, levels = c("Uber", "Lyft")),
    time_hour = reorder(time_hour, time_hour, FUN = sum),
    week_day = factor(Weekdays[week_day], levels = Weekdays)
  )][,
    better_than_model := median(daily_hourly_wage_mean) > FinalPfaMeans[2L],
    by = c("time_hour", "taxi_company", "week_day")
  ],
  aes(
    time_hour,
    daily_hourly_wage_mean,
    fill = better_than_model
  )
) +
  geom_boxplot() +
  geom_hline(yintercept = FinalPfaMeans[1L], linetype = 2) +
  geom_hline(yintercept = FinalPfaMeans[2L], linetype = 2) +
  scale_fill_manual(
    values = BetterThanModelFillColors
  ) +
  scale_y_continuous(breaks = BreaksToPlot, label = scales::dollar_format(1)) +
  facet_grid(vars(taxi_company), vars(week_day)) +
  labs(
    title = paste0(
      "<span style='color:",
      Params$ColorHighlight,
      ";'>**Start Time**</span> ",
      "**Can Increase Earning Reguardless Day of Week**"
    ),
    x = "Hour Of The Day",
    y = "Daily Hourly Wage"
  ) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(
    plot.title = element_markdown(size = 14, family = Params$BaseFontFamily),
    plot.subtitle = element_markdown(size = 12, family = Params$BaseFontFamily),
    axis.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none",
    strip.text = element_text(
      size = 12,
      face = "bold",
      family = Params$BaseFontFamily
    )
  )
```

## Modeling Exploration

Based on the prior exploration it's clear that the most important rule found so far is to take Uber as the company, but to improve earning we need to find better interactions between variables and for that reason we are going to start with tree model to see the model can find the most important interactions to explain changes in the target variable.

### Tree models

To get the best from the tree model we will tune the following params:

* `tree_depth` to limit how many splits the tree can make (controls vertical complexity).
* `min_n` to set the minimum number of data points required in a node to allow a further split (controls leaf size and horizontal complexity).

For the feature engeniereeng part we took the following decisions:

* `start_zone` to about the problem of high cardinality of zone we substitute the each zone by their mean after separaring the effect of the company.
* `taxi_company` was transformed into binary varible as we only have 2 companies for know.
* `time_hour` was transform to make splict the 24 hour cycle we face everyday.
* `week_day` was combined with the `time_hour` to make explicit the cycle that we face every week.

```{r}
#| label: Defining Tree Workdflow
DecisionTreeSpec <- decision_tree(
  mode = "classification",
  tree_depth = tune("tree_depth"),
  min_n = tune("min_n_trees")
)

ExperimentTreeRecipe <- recipe(
  is_top_trip ~ taxi_company +
    start_zone +
    week_day +
    time_hour +
    simulation_id +
    daily_hourly_wage_mean,
  data = InitialZoneSimResults
) |>
  update_role(
    simulation_id,
    daily_hourly_wage_mean,
    new_role = "additional info"
  ) |>
  # Creating variables to use later
  step_mutate(
    week_time = (week_day - 1) * 24 + time_hour,
    start_zone_taxi = paste0(start_zone, "-", taxi_company)
  ) |>
  # Changing zones with their mean
  step_lencode(
    start_zone_taxi,
    outcome = vars(daily_hourly_wage_mean),
    smooth = FALSE
  ) |>
  # Transforming company to binary variable
  step_dummy(taxi_company) |>
  # Making clear 24h cycles
  step_harmonic(
    time_hour,
    cycle_size = 24,
    frequency = 1,
    keep_original_cols = TRUE
  ) |>
  step_harmonic(
    week_time,
    cycle_size = 168,
    frequency = 1,
    keep_original_cols = TRUE
  ) |>
  # Removing original variables
  step_rm(week_day, time_hour, start_zone, week_time)

ExperimentTreeRecipeBalance <- step_upsample(ExperimentTreeRecipe, is_top_trip)

ExperimentTreeWfSet <- workflow_set(
  preproc = list(
    TreeTransformations = ExperimentTreeRecipe,
    TreeTransformationsBalance = ExperimentTreeRecipeBalance
  ),
  models = list(
    tree = DecisionTreeSpec
  )
)
```

As this as simple model and we just need to tune 2 params we can use the grid method the find the best trip possible with 10 fold cross validation to estimate the model error without taking testing data, as we just need to understand the results of the experiment rather than using this model to make predictions.

```{r}
#| eval: false
#| label: Tuning tree model
ExperimentTreeWfRegularGrid <-
  extract_parameter_set_dials(
    ExperimentTreeWfSet,
    id = "TreeTransformations_tree"
  ) |>
  grid_regular(levels = 5)

set.seed(5878)
InitialZoneSimResultsResamples <- vfold_cv(
  InitialZoneSimResults,
  v = 10,
  strata = is_top_trip
)

# Tunning and saving ----
ExperimentTreeWfTuned <- list()

for (wf_id_i in ExperimentTreeWfSet$wflow_id) {
  wf_i <- extract_workflow(ExperimentTreeWfSet, id = wf_id_i)
  cat(paste("Tuning", wf_id_i))

  ExperimentTreeWfTuned[[wf_id_i]] <- tune_grid(
    object = wf_i,
    resamples = InitialZoneSimResultsResamples,
    grid = ExperimentTreeWfRegularGrid,
    metrics = metric_set(roc_auc, brier_class),
    control = control_grid(
      verbose = TRUE,
      parallel_over = NULL,
      save_pred = TRUE
    )
  )

  cat(paste("Completed", wf_id_i))

  # Free memory
  gc()
}

pin_write(
  BoardLocal,
  ExperimentTreeWfTuned,
  name = "ExperimentTreeWfTuned",
  type = "qs2"
)
```

Based on the results we can see that `min_n_trees` don't present any change on model performance, that we get the best results when `min_n` is at lest 4.

```{r}
ExperimentTreeWfTuned <- pin_read(
  BoardLocal,
  "ExperimentTreeWfTuned"
)

ExperimentTreeWfBestConfig <- select_best(
  ExperimentTreeWfTuned$TreeTransformationsBalance_tree,
  metric = "brier_class"
)

autoplot(ExperimentTreeWfTuned$TreeTransformations_tree) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(legend.position = "top")

autoplot(ExperimentTreeWfTuned$TreeTransformationsBalance_tree) +
  theme_light(base_family = Params$BaseFontFamily) +
  theme(legend.position = "top")
```

```{r}
#| eval: false
collect_predictions(ExperimentTreeWfTuned$TreeTransformationsBalance_tree) |>
  semi_join(ExperimentTreeWfBestConfig, by = ".config")
```


```{r}
#| eval: false
ExperimentTreeWfFitted <- ExperimentTreeWfSet |>
  extract_workflow(id = "TreeTransformationsBalance_tree") |>
  finalize_workflow(parameters = ExperimentTreeWfBestConfig) |>
  fit(data = InitialZoneSimResults)

pin_write(
  BoardLocal,
  ExperimentTreeWfFitted,
  name = "ExperimentTreeWfFitted",
  type = "qs2"
)
```

Based on the model results we can see that the `start_zone_taxi` became the must important variable to start the model then we see how

```{r}
ExperimentTreeWfFitted <- pin_read(
  BoardLocal,
  "ExperimentTreeWfFitted"
)

collect_metrics(ExperimentTreeWfTuned$TreeTransformationsBalance_tree) |>
  semi_join(ExperimentTreeWfBestConfig, by = ".config")

extract_fit_engine(ExperimentTreeWfFitted)
```


```{r}
# El árbol nos dijo 3 cosas clave:
# 1. start_zone_taxi es el predictor dominante (threshold ~$59.65)
# 2. time_hour tiene efecto armónico (sin < 0 = primera mitad del día)
# 3. Interacción hora × día_semana (fin de semana nocturno es especial)

LinearRecipe <- recipe(
  is_top_trip ~ taxi_company +
    start_zone +
    week_day +
    time_hour +
    simulation_id +
    daily_hourly_wage_mean,
  data = InitialZoneSimResults
) |>
  update_role(
    simulation_id,
    daily_hourly_wage_mean,
    new_role = "additional info"
  ) |>

  # --- Igual que el árbol ---
  step_mutate(
    week_time = (week_day - 1) * 24 + time_hour,
    start_zone_taxi = paste0(start_zone, "-", taxi_company)
  ) |>
  step_lencode(
    # mixed mejor que bayes para regresión logística
    start_zone_taxi,
    outcome = vars(daily_hourly_wage_mean),
    smooth = FALSE # shrinkage evita overfitting en zonas con pocas obs
  ) |>
  step_dummy(taxi_company) |>
  step_harmonic(
    time_hour,
    cycle_size = 24,
    frequency = 1,
    keep_original_cols = TRUE
  ) |>
  step_harmonic(
    week_time,
    cycle_size = 168,
    frequency = 1,
    keep_original_cols = TRUE
  ) |>

  # --- Nuevo: lo que el árbol descubrió como interacciones ---

  # Threshold de zona (~$59.65): captura el salto que encontró el árbol
  step_mutate(
    zona_top = as.integer(start_zone_taxi >= 59.65)
  ) |>

  # Interacción 1: zona × hora (el árbol dijo que hora solo importa en zonas buenas)
  step_interact(
    terms = ~ zona_top:time_hour_sin_1 + zona_top:time_hour_cos_1
  ) |>

  # Interacción 2: zona × fin de semana nocturno (el tercer split del árbol)
  step_interact(
    terms = ~ zona_top:week_time_cos_1 + zona_top:week_time_sin_1
  ) |>

  step_rm(week_day, time_hour, start_zone, week_time, zona_top) |>
  step_zv(all_predictors()) |> # elimina constantes post-interacción
  step_normalize(all_numeric_predictors()) # necesario para logística

LogisticSpec <- logistic_reg() |>
  set_engine("glm")

LogisticWf <- workflow() |>
  add_recipe(LinearRecipe) |>
  add_model(LogisticSpec)
```

```{r}
#| eval: false
set.seed(5878)
InitialZoneSimResultsResamples <- vfold_cv(
  InitialZoneSimResults,
  v = 10,
  strata = is_top_trip
)

# Reusar los mismos folds del árbol para comparación justa
LogisticWfResampled <- fit_resamples(
  LogisticWf,
  resamples = InitialZoneSimResultsResamples,
  metrics = metric_set(roc_auc, brier_class),
  control = control_resamples(
    verbose = TRUE,
    save_pred = TRUE
  )
)

pin_write(
  BoardLocal,
  LogisticWfResampled,
  name = "LogisticWfResampled",
  type = "qs2"
)
```

```{r}
LogisticWfResampled <- pin_read(
  BoardLocal,
  name = "LogisticWfResampled"
)

# Métricas lado a lado
tree_metrics <- collect_metrics(
  ExperimentTreeWfTuned$TreeTransformationsBalance_tree
) |>
  semi_join(ExperimentTreeWfBestConfig, by = ".config") |>
  mutate(model = "Decision Tree")

logistic_metrics <- collect_metrics(LogisticWfResampled) |>
  mutate(model = "Logistic Regression")

bind_rows(tree_metrics, logistic_metrics) |>
  select(model, .metric, mean, std_err) |>
  arrange(.metric, model)
```

```{r}
LogisticWfFitted <- fit(LogisticWf, data = InitialZoneSimResults)

# Para ver significancia de cada feature
LogisticWfFitted |>
  extract_fit_engine() |>
  summary()
```

```{r}
# Para comparar con el árbol en mismo framework
LogisticWfFitted |>
  extract_fit_engine() |>
  broom::tidy(conf.int = TRUE) |>
  filter(p.value < 0.05) |>
  arrange(desc(abs(estimate)))
```

