---
title: "From Predictions to Policies: Integrating ML into Stochastic Optimization"
subtitle: ""
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once have a policy to decide if a trip should be taken or rejected then it's important to validate if any of the next factors could cause an increase of decrease on taxi results after blocking the **company** for taxi driver:

- Start zone
- Days of the week
- Hour of the day

## Defining combinations to simulate

As we have 196 zones to evaluete, 3 days to evaluete for each day of the week (7), 2 taxi companies and we divide the day in 8 moments to measure (one each 3 hours), we end with an experiment of **65,856 days** to simulate.

```{r}
#| label: Defining combinations
# Loading functions ----

## To manage relative paths
library(here)
library(pins)
library(qs2)

## Manipulate data
library(data.table)
library(DBI)
library(duckdb)
library(lubridate)
library(timeDate)

## To run predictions
library(tidymodels)

## Custom functions, board & params
devtools::load_all()
BoardLocal <- board_folder(here("../NycTaxiPins/Board"))
Params <- yaml::read_yaml(here("params.yml"))


# Defining combinations ----

## Getting shapes to use
ZonesShapes <-
  pin_read(BoardLocal, "ZonesShapes") |>
  subset(borough %in% names(Params$BoroughColors))

UniqueZoneIds <- as.character(ZonesShapes$LocationID) |> unique() |> sort()


## Defining days to sample
YearDays <- data.table(
  days = seq.Date(make_date(2023), make_date(2023, 12, 31), by = "day")
)
YearDays[, wday := lubridate::wday(days)]
set.seed(4879)
SampledDays <- YearDays[, .SD[sample.int(.N, 3L)], by = "wday"]


## Pre-build the big design matrix once
DataToSimulate <- CJ(
  DOLocationID = UniqueZoneIds,
  hvfhs_license_num = c("HV0003", "HV0005"),
  calendar_day = as.character(SampledDays$days),
  time_hour = seq(0, 21, length.out = 8)
)
DataToSimulate[, `:=`(
  trip_id = as.numeric(1:.N),
  request_datetime = paste(calendar_day, time_hour) |> ymd_h(),
  trip_time = 0
)]

head(DataToSimulate, 8)
```

## Running simulation

As I knew that running the hole process would take me several days I decided to store the results of simulations on disk for each zone. In this way I was able to keep the experiment as I was expecting and stop the process between days to allow the computer to rest between days and redice the risk of loossing all the progress I had problems with memory while using the computer for dealy activities.

To avoid the overhead of using a IDE the next code was source from terminal after connecting with `ssh NycTaxi` to consume the fewer results as possible.

::: {.callout-note}
In the below code you see an assigning the starting zone to `DOLocationID` as the simulation uses the last zone of the configuration trip (the one we are creting) to starting picking trips from that zone.
:::

```{r}
#| eval: false
#| label: Simulation code
## Loading env

# Getting prior policy
AcceptRejectPolicyFitted <-
  pin_read(BoardLocal, "AcceptRejectPolicyFitted")

# Competing renaming columns
DataToSimulate[, `:=`(
  calendar_day = NULL,
  time_hour = NULL,
  PULocationID = "1",
  wav_match_flag = "N",
  driver_pay = 0,
  tips = 0,
  dispatching_base_num = "",
  originating_base_num = "",
  on_scene_datetime = NA_POSIXct_,
  pickup_datetime = NA_POSIXct_,
  dropoff_datetime = NA_POSIXct_,
  trip_miles = 0,
  base_passenger_fare = 0,
  tolls = 0,
  bcf = 0,
  sales_tax = 0,
  congestion_surcharge = 0,
  airport_fee = 0,
  shared_request_flag = "",
  shared_match_flag = "",
  access_a_ride_flag = "",
  wav_request_flag = "",
  month = 1,
  year = 2023,
  performance_per_hour = 0
)]

# Check which zones are already done
AlreadyDone <- pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  gsub(pattern = "InitialZoneSim_ZoneId_", replacement = "")

zones_to_run <- setdiff(UniqueZoneIds, AlreadyDone)

cat(
  "→",
  length(zones_to_run),
  "zones left to simulate (out of",
  length(UniqueZoneIds),
  "total).\n"
)


# === MAIN RESUMABLE LOOP ===
for (zone_i in zones_to_run) {
  cat("   Simulating zone", zone_i, "... ")

  # Filter only this zone
  ZoneData <- DataToSimulate[list(zone_i), on = "DOLocationID"]

  # Fresh connection for each zone
  con <- dbConnect(duckdb(), dbdir = here("../NycTaxiBigFiles/my-db.duckdb"))

  ZoneSim <- simulate_trips(
    con,
    ZoneData,
    seeds = 4878,
    fitted_wf = AcceptRejectPolicyFitted
  )

  dbDisconnect(con, shutdown = TRUE)

  # Save immediately with clear name
  pin_write(
    BoardLocal,
    ZoneSim,
    paste0("InitialZoneSim_ZoneId_", zone_i),
    type = "qs2"
  )

  cat("DONE (", nrow(ZoneSim), "rows)\n")
}

cat("✅ All zones simulated!\n")
```

Here is the list of zone simulated and we can see that it took around **102 minutes** to complete each zone which mean that I invest around **14 days** of computation to see the next results.

```bash
angelfeliz@debian:~/r-projects/NycTaxi$ ls -lahd ../NycTaxiPins/Board/InitialZoneSim_ZoneId_*
drwxr-xr-x 3 root root 4.0K Apr  7 10:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_10
drwxr-xr-x 3 root root 4.0K Apr  7 12:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_100
drwxr-xr-x 3 root root 4.0K Apr  7 13:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_101
drwxr-xr-x 3 root root 4.0K Apr  7 15:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_102
drwxr-xr-x 3 root root 4.0K Apr  7 17:06 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_103
drwxr-xr-x 3 root root 4.0K Apr  7 18:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_106
drwxr-xr-x 3 root root 4.0K Apr  7 20:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_107
drwxr-xr-x 3 root root 4.0K Apr  7 22:31 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_108
drwxr-xr-x 3 root root 4.0K Apr  8 00:11 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_11
drwxr-xr-x 3 root root 4.0K Apr  8 01:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_111
drwxr-xr-x 3 root root 4.0K Apr  8 03:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_112
drwxr-xr-x 3 root root 4.0K Apr  8 05:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_113
drwxr-xr-x 3 root root 4.0K Apr  8 06:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_114
drwxr-xr-x 3 root root 4.0K Apr  8 08:35 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_116
drwxr-xr-x 3 root root 4.0K Apr  8 10:16 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_117
drwxr-xr-x 3 root root 4.0K Apr  8 11:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_12
drwxr-xr-x 3 root root 4.0K Apr  8 13:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_120
drwxr-xr-x 3 root root 4.0K Apr  8 15:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_121
drwxr-xr-x 3 root root 4.0K Apr  8 16:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_122
drwxr-xr-x 3 root root 4.0K Apr  8 18:40 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_123
drwxr-xr-x 3 root root 4.0K Apr  8 20:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_124
drwxr-xr-x 3 root root 4.0K Apr  8 22:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_125
drwxr-xr-x 3 root root 4.0K Apr  9 01:50 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_127
drwxr-xr-x 3 root root 4.0K Apr  9 03:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_128
drwxr-xr-x 3 root root 4.0K Apr  9 05:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_129
drwxr-xr-x 3 root root 4.0K Apr  9 06:51 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_13
drwxr-xr-x 3 root root 4.0K Apr  9 08:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_130
drwxr-xr-x 3 root root 4.0K Apr  9 11:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_131
drwxr-xr-x 3 root root 4.0K Apr  9 13:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_132
drwxr-xr-x 3 root root 4.0K Apr  9 15:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_133
drwxr-xr-x 3 root root 4.0K Apr  9 16:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_134
drwxr-xr-x 3 root root 4.0K Apr  9 18:23 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_135
drwxr-xr-x 3 root root 4.0K Apr  9 20:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_137
drwxr-xr-x 3 root root 4.0K Apr  9 21:51 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_138
drwxr-xr-x 3 root root 4.0K Apr  9 23:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_139
drwxr-xr-x 3 root root 4.0K Apr 10 01:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_14
drwxr-xr-x 3 root root 4.0K Apr 10 03:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_140
drwxr-xr-x 3 root root 4.0K Apr 10 05:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_141
drwxr-xr-x 3 root root 4.0K Apr 10 06:41 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_142
drwxr-xr-x 3 root root 4.0K Apr 10 10:25 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_143
drwxr-xr-x 3 root root 4.0K Apr 10 12:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_144
drwxr-xr-x 3 root root 4.0K Apr 10 13:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_145
drwxr-xr-x 3 root root 4.0K Apr 10 15:33 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_146
drwxr-xr-x 3 root root 4.0K Apr 10 17:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_148
drwxr-xr-x 3 root root 4.0K Apr 10 18:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_149
drwxr-xr-x 3 root root 4.0K Apr 10 20:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_15
drwxr-xr-x 3 root root 4.0K Apr 10 22:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_150
drwxr-xr-x 3 root root 4.0K Apr 11 00:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_151
drwxr-xr-x 3 root root 4.0K Apr 11 01:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_152
drwxr-xr-x 3 root root 4.0K Apr 11 03:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_153
drwxr-xr-x 3 root root 4.0K Apr 11 05:02 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_154
drwxr-xr-x 3 root root 4.0K Apr 11 06:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_155
drwxr-xr-x 3 root root 4.0K Apr 11 08:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_157
drwxr-xr-x 3 root root 4.0K Apr 11 10:07 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_158
drwxr-xr-x 3 root root 4.0K Apr 11 11:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_16
drwxr-xr-x 3 root root 4.0K Apr 11 13:47 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_160
drwxr-xr-x 3 root root 4.0K Apr 11 15:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_161
drwxr-xr-x 3 root root 4.0K Apr 12 00:48 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_162
drwxr-xr-x 3 root root 4.0K Apr 12 02:32 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_163
drwxr-xr-x 3 root root 4.0K Apr 12 04:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_164
drwxr-xr-x 3 root root 4.0K Apr 12 05:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_165
drwxr-xr-x 3 root root 4.0K Apr 12 07:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_166
drwxr-xr-x 3 root root 4.0K Apr 12 09:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_17
drwxr-xr-x 3 root root 4.0K Apr 12 10:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_170
drwxr-xr-x 3 root root 4.0K Apr 12 12:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_171
drwxr-xr-x 3 root root 4.0K Apr 12 14:21 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_173
drwxr-xr-x 3 root root 4.0K Apr 12 16:13 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_175
drwxr-xr-x 3 root root 4.0K Apr 12 18:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_177
drwxr-xr-x 3 root root 4.0K Apr 12 20:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_178
drwxr-xr-x 3 root root 4.0K Apr 12 22:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_179
drwxr-xr-x 3 root root 4.0K Apr 12 23:49 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_180
drwxr-xr-x 3 root root 4.0K Apr 13 01:33 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_181
drwxr-xr-x 3 root root 4.0K Apr 13 03:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_186
drwxr-xr-x 3 root root 4.0K Apr 13 04:56 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_188
drwxr-xr-x 3 root root 4.0K Apr 13 06:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_189
drwxr-xr-x 3 root root 4.0K Apr 13 08:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_19
drwxr-xr-x 3 root root 4.0K Apr 13 09:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_190
drwxr-xr-x 3 root root 4.0K Apr 13 11:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_191
drwxr-xr-x 3 root root 4.0K Apr 13 13:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_192
drwxr-xr-x 3 root root 4.0K Apr 13 14:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_193
drwxr-xr-x 3 root root 4.0K Apr 13 18:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_194
drwxr-xr-x 3 root root 4.0K Apr 13 19:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_195
drwxr-xr-x 3 root root 4.0K Apr 13 21:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_196
drwxr-xr-x 3 root root 4.0K Apr 13 23:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_197
drwxr-xr-x 3 root root 4.0K Apr 14 01:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_198
drwxr-xr-x 3 root root 4.0K Apr 14 02:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_2
drwxr-xr-x 3 root root 4.0K Apr 14 04:35 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_201
drwxr-xr-x 3 root root 4.0K Apr 14 06:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_202
drwxr-xr-x 3 root root 4.0K Apr 14 07:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_203
drwxr-xr-x 3 root root 4.0K Apr 14 09:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_205
drwxr-xr-x 3 root root 4.0K Apr 14 11:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_207
drwxr-xr-x 3 root root 4.0K Apr 14 13:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_209
drwxr-xr-x 3 root root 4.0K Apr 14 14:44 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_21
drwxr-xr-x 3 root root 4.0K Apr 14 16:27 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_210
drwxr-xr-x 3 root root 4.0K Apr 14 18:09 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_211
drwxr-xr-x 3 root root 4.0K Apr 14 22:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_215
drwxr-xr-x 3 root root 4.0K Apr 15 00:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_216
drwxr-xr-x 3 root root 4.0K Apr 15 02:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_217
drwxr-xr-x 3 root root 4.0K Apr 15 04:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_218
drwxr-xr-x 3 root root 4.0K Apr 15 05:56 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_219
drwxr-xr-x 3 root root 4.0K Apr 15 07:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_22
drwxr-xr-x 3 root root 4.0K Apr 15 09:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_222
drwxr-xr-x 3 root root 4.0K Apr 15 11:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_223
drwxr-xr-x 3 root root 4.0K Apr 15 13:02 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_224
drwxr-xr-x 3 root root 4.0K Apr 15 14:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_225
drwxr-xr-x 3 root root 4.0K Apr 15 16:28 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_226
drwxr-xr-x 3 root root 4.0K Apr 15 18:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_227
drwxr-xr-x 3 root root 4.0K Apr 15 19:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_228
drwxr-xr-x 3 root root 4.0K Apr 15 21:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_229
drwxr-xr-x 3 root root 4.0K Apr 15 23:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_230
drwxr-xr-x 3 root root 4.0K Apr 16 00:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_231
drwxr-xr-x 3 root root 4.0K Apr 16 02:38 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_232
drwxr-xr-x 3 root root 4.0K Apr 16 04:18 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_233
drwxr-xr-x 3 root root 4.0K Apr 16 05:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_234
drwxr-xr-x 3 root root 4.0K Apr 16 07:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_236
drwxr-xr-x 3 root root 4.0K Apr 16 09:22 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_237
drwxr-xr-x 3 root root 4.0K Apr 16 11:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_238
drwxr-xr-x 3 root root 4.0K Apr 16 12:45 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_239
drwxr-xr-x 3 root root 4.0K Apr 16 14:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_24
drwxr-xr-x 3 root root 4.0K Apr 16 16:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_243
drwxr-xr-x 3 root root 4.0K Apr 16 17:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_244
drwxr-xr-x 3 root root 4.0K Apr 16 19:25 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_246
drwxr-xr-x 3 root root 4.0K Apr 16 21:08 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_249
drwxr-xr-x 3 root root 4.0K Apr 16 22:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_25
drwxr-xr-x 3 root root 4.0K Apr 17 00:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_252
drwxr-xr-x 3 root root 4.0K Apr 17 02:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_253
drwxr-xr-x 3 root root 4.0K Apr 17 04:17 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_255
drwxr-xr-x 3 root root 4.0K Apr 17 05:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_256
drwxr-xr-x 3 root root 4.0K Apr 17 07:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_257
drwxr-xr-x 3 root root 4.0K Apr 17 09:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_258
drwxr-xr-x 3 root root 4.0K Apr 17 11:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_26
drwxr-xr-x 3 root root 4.0K Apr 17 13:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_260
drwxr-xr-x 3 root root 4.0K Apr 17 14:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_261
drwxr-xr-x 3 root root 4.0K Apr 17 16:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_262
drwxr-xr-x 3 root root 4.0K Apr 17 18:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_263
drwxr-xr-x 3 root root 4.0K Apr 17 19:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_27
drwxr-xr-x 3 root root 4.0K Apr 17 21:37 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_28
drwxr-xr-x 3 root root 4.0K Apr 17 23:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_29
drwxr-xr-x 3 root root 4.0K Apr 18 00:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_30
drwxr-xr-x 3 root root 4.0K Apr 18 02:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_33
drwxr-xr-x 3 root root 4.0K Apr 18 04:16 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_34
drwxr-xr-x 3 root root 4.0K Apr 18 05:55 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_35
drwxr-xr-x 3 root root 4.0K Apr 18 07:40 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_36
drwxr-xr-x 3 root root 4.0K Apr 18 09:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_37
drwxr-xr-x 3 root root 4.0K Apr 18 11:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_38
drwxr-xr-x 3 root root 4.0K Apr 18 13:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_39
drwxr-xr-x 3 root root 4.0K Apr 18 15:03 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_4
drwxr-xr-x 3 root root 4.0K Apr 18 16:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_40
drwxr-xr-x 3 root root 4.0K Apr 18 19:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_41
drwxr-xr-x 3 root root 4.0K Apr 18 20:52 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_42
drwxr-xr-x 3 root root 4.0K Apr 18 22:41 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_43
drwxr-xr-x 3 root root 4.0K Apr 19 00:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_45
drwxr-xr-x 3 root root 4.0K Apr 19 02:15 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_48
drwxr-xr-x 3 root root 4.0K Apr 19 03:57 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_49
drwxr-xr-x 3 root root 4.0K Apr 19 05:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_50
drwxr-xr-x 3 root root 4.0K Apr 19 07:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_52
drwxr-xr-x 3 root root 4.0K Apr 19 09:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_53
drwxr-xr-x 3 root root 4.0K Apr 19 10:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_54
drwxr-xr-x 3 root root 4.0K Apr 19 12:29 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_55
drwxr-xr-x 3 root root 4.0K Apr 19 14:11 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_56
drwxr-xr-x 3 root root 4.0K Apr 19 16:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_61
drwxr-xr-x 3 root root 4.0K Apr 19 17:50 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_62
drwxr-xr-x 3 root root 4.0K Apr 19 19:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_63
drwxr-xr-x 3 root root 4.0K Apr 19 21:27 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_64
drwxr-xr-x 3 root root 4.0K Apr 19 23:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_65
drwxr-xr-x 3 root root 4.0K Apr 20 01:01 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_66
drwxr-xr-x 3 root root 4.0K Apr 20 02:43 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_67
drwxr-xr-x 3 root root 4.0K Apr 20 04:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_68
drwxr-xr-x 3 root root 4.0K Apr 20 06:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_7
drwxr-xr-x 3 root root 4.0K Apr 20 07:46 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_70
drwxr-xr-x 3 root root 4.0K Apr 20 09:31 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_71
drwxr-xr-x 3 root root 4.0K Apr 20 11:10 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_72
drwxr-xr-x 3 root root 4.0K Apr 20 12:49 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_73
drwxr-xr-x 3 root root 4.0K Apr 20 14:34 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_74
drwxr-xr-x 3 root root 4.0K Apr 20 16:20 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_75
drwxr-xr-x 3 root root 4.0K Apr 20 18:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_76
drwxr-xr-x 3 root root 4.0K Apr 20 22:13 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_77
drwxr-xr-x 3 root root 4.0K Apr 21 00:36 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_79
drwxr-xr-x 3 root root 4.0K Apr 21 02:54 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_8
drwxr-xr-x 3 root root 4.0K Apr 21 09:47 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_80
drwxr-xr-x 3 root root 4.0K Apr 21 11:30 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_82
drwxr-xr-x 3 root root 4.0K Apr 21 13:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_83
drwxr-xr-x 3 root root 4.0K Apr 21 17:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_85
drwxr-xr-x 3 root root 4.0K Apr 21 18:58 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_86
drwxr-xr-x 3 root root 4.0K Apr 21 20:59 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_87
drwxr-xr-x 3 root root 4.0K Apr 21 23:00 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_88
drwxr-xr-x 3 root root 4.0K Apr 22 00:45 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_89
drwxr-xr-x 3 root root 4.0K Apr 22 02:24 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_9
drwxr-xr-x 3 root root 4.0K Apr 22 04:04 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_90
drwxr-xr-x 3 root root 4.0K Apr 22 05:44 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_91
drwxr-xr-x 3 root root 4.0K Apr 22 07:42 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_92
drwxr-xr-x 3 root root 4.0K Apr 22 09:26 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_93
drwxr-xr-x 3 root root 4.0K Apr 22 11:14 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_95
drwxr-xr-x 3 root root 4.0K Apr 22 12:53 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_96
drwxr-xr-x 3 root root 4.0K Apr 22 14:39 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_97
drwxr-xr-x 3 root root 4.0K Apr 22 16:19 ../NycTaxiPins/Board/InitialZoneSim_ZoneId_98
```

## Exploring results

```{r}
#| eval: false
#| label: Consolidating results
consolidate_all_simulations <- function(zone_id_data, board, sim_start_days) {
  consolidated_data =
    sim_start_trip_summary(
      pin_read(board, zone_id_data),
      sim_start_days
    )[,
      sim_start_days[, list(
        simulation_id = trip_id,
        taxi_company = fifelse(hvfhs_license_num == 'HV0003', "Uber", "Lyft"),
        start_zone = DOLocationID,
        week_day = wday(calendar_day),
        time_hour
      )][.SD, on = "simulation_id"],
      .SDcols = !patterns("_sd")
    ]

  return(consolidated_data)
}

InitialZoneSimResults <-
  pin_list(BoardLocal) |>
  grep(pattern = "^InitialZoneSim_ZoneId_", value = TRUE) |>
  lapply(
    FUN = consolidate_all_simulations,
    board = BoardLocal,
    sim_start_days = DataToSimulate
  ) |>
  rbindlist()

# Save immediately with clear name
pin_write(
  BoardLocal,
  InitialZoneSimResults,
  "InitialZoneSimResults",
  type = "qs2"
)
```


```{r}
pin_read(
  BoardLocal,
  "InitialZoneSimResults"
)[, .N, keyby = c("week_day", "time_hour")]


# Assuming your data is called 'sim_results'
rec <- recipe(~ week_day + time_hour, data = sim_results) %>%
  # 1. Combine day and hour into a single timeline (0 to 167)
  step_mutate(
    week_time = (week_day - 1) * 24 + time_hour
  ) %>%
  # 2. Create sine and cosine features based on a 168-hour cycle
  step_harmonic(
    week_time,
    cycle_size = 168,
    frequency = 1,
    keep_original_cols = FALSE
  )

# Prepare and bake to see the results
processed_data <- prep(rec) %>% bake(new_data = NULL)
```